---
title: "Homework 4 DATA-624"
author: "Mehreen Ali Gillani"
date: "2026-02-26"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Load required libraries
library(ggplot2)
library(corrplot)
library(gridExtra)
library(tidyr)
library(MASS)
library(moments)
par(mar = c(5, 4, 4, 2) + 0.1)
par(mfrow = c(1, 1))
par(oma = c(0, 0, 0, 0))
par(ask = FALSE)
par(bg = "white")



```

3.1. The UC Irvine Machine Learning Repository 6 contains a dataset
related to glass identification. The data consist of 214 glass samples
labeled as one of seven class categories. There are nine predictors,
including the refractive index and percentages of eight elements: Na,
Mg, Al, Si, K, Ca, Ba, and Fe. The data can be accessed via:

```{r}
# Load the dataset
library(mlbench)
data(Glass)
str(Glass)
```

a)  Using visualizations, explore the predictor variables to understand
    their distributions as well as the relationships between predictors.

```{r}

# 1. Distribution plots for each predictor
# Create histograms with density curves
predictors <- names(Glass)[1:9]  # Excluding Type (response variable)

# Histograms
hist_plots <- lapply(predictors, function(var) {
  ggplot(Glass, aes_string(x = var)) +
    geom_histogram(aes(y = ..density..), bins = 20, 
                   fill = "steelblue", color = "black", alpha = 0.7) +
    geom_density(color = "red", size = 1) +
    theme_minimal() +
    labs(title = paste("Distribution of", var))
})

# Arrange histograms in a grid
do.call(grid.arrange, c(hist_plots, ncol = 3))

# 2. Boxplots to see distributions and outliers
boxplot_data <- Glass[, predictors] %>%
  gather(key = "variable", value = "value")

ggplot(boxplot_data, aes(x = variable, y = value)) +
  geom_boxplot(fill = "steelblue", alpha = 0.7) +
  theme_minimal() +
  labs(title = "Boxplots of Predictor Variables") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# 3. Correlation matrix to understand relationships
cor_matrix <- cor(Glass[, predictors])
corrplot(cor_matrix, method = "color", type = "upper", 
         order = "hclust", tl.cex = 0.8, 
         title = "Correlation Matrix of Predictors",
         mar = c(0, 0, 2, 0))

# 4. Scatterplot matrix (simplified version)
pairs(Glass[, predictors], pch = 19, cex = 0.5, 
      col = rgb(0, 0, 1, 0.5), 
      main = "Scatterplot Matrix of Predictors")

# 5. Relationship with target variable (Type)
# Boxplots by Type for each predictor
type_plots <- lapply(predictors[1:6], function(var) {  # First 6 predictors for readability
  ggplot(Glass, aes_string(x = "Type", y = var, fill = "Type")) +
    geom_boxplot(alpha = 0.7) +
    theme_minimal() +
    theme(legend.position = "none") +
    labs(title = paste(var, "by Glass Type"))
})

do.call(grid.arrange, c(type_plots, ncol = 2))

# 6. Summary statistics
summary(Glass[, predictors])
```

##### (b) Do there appear to be any outliers in the data? Are any predictors skewed?

##### Outlier Detection

```{r}

# Corrected boxplots with jitter
outlier_plots <- lapply(predictors, function(var) {
  ggplot(Glass, aes_string(x = factor(0), y = var)) +  # Add dummy x variable
    geom_boxplot(fill = "steelblue", alpha = 0.7, width = 0.5) +
    geom_jitter(width = 0.2, alpha = 0.3, color = "red") +
    theme_minimal() +
    labs(title = var, x = "") +
    theme(axis.text.x = element_blank(),
          axis.ticks.x = element_blank())
})

# Display plots
do.call(grid.arrange, c(outlier_plots, ncol = 3, 
                        top = "Boxplots with Outliers Highlighted"))

# Alternative: simpler boxplots without jitter
simple_boxplots <- lapply(predictors, function(var) {
  ggplot(Glass, aes_string(y = var)) +
    geom_boxplot(fill = "steelblue", alpha = 0.7) +
    theme_minimal() +
    labs(title = var, y = var) +
    theme(axis.text.x = element_blank(),
          axis.ticks.x = element_blank())
})

do.call(grid.arrange, c(simple_boxplots, ncol = 3, 
                        top = "Boxplots of Predictors"))

# Quantitative outlier detection using IQR method
detect_outliers <- function(x) {
  Q1 <- quantile(x, 0.25, na.rm = TRUE)
  Q3 <- quantile(x, 0.75, na.rm = TRUE)
  IQR <- Q3 - Q1
  lower_bound <- Q1 - 1.5 * IQR
  upper_bound <- Q3 + 1.5 * IQR
  outliers <- sum(x < lower_bound | x > upper_bound, na.rm = TRUE)
  return(outliers)
}

outlier_counts <- sapply(Glass[, predictors], detect_outliers)
outlier_df <- data.frame(
  Variable = names(outlier_counts),
  Outliers = outlier_counts,
  Percentage = round(outlier_counts / nrow(Glass) * 100, 2)
)
print("Outlier Counts:")
print(outlier_df[order(-outlier_df$Outliers), ])

# Calculate skewness
skewness_values <- sapply(Glass[, predictors], skewness)
skewness_df <- data.frame(
  Variable = names(skewness_values),
  Skewness = round(skewness_values, 3),
  Skewness_Type = ifelse(abs(skewness_values) < 0.5, "Symmetric",
                         ifelse(abs(skewness_values) < 1, "Moderately Skewed", 
                                "Highly Skewed"))
)
print("\nSkewness Analysis:")
print(skewness_df[order(-abs(skewness_values)), ])

# Corrected histograms with statistics
hist_with_stats <- lapply(predictors, function(var) {
  skew_val <- round(skewness(Glass[[var]]), 2)
  mean_val <- round(mean(Glass[[var]]), 2)
  median_val <- round(median(Glass[[var]]), 2)
  
  ggplot(Glass, aes_string(x = var)) +
    geom_histogram(aes(y = ..density..), bins = 20, 
                   fill = "steelblue", color = "black", alpha = 0.7) +
    geom_density(color = "red", size = 1) +
    geom_vline(aes(xintercept = mean_val), 
               color = "green", linetype = "dashed", size = 1) +
    geom_vline(aes(xintercept = median_val), 
               color = "orange", linetype = "dashed", size = 1) +
    annotate("text", x = Inf, y = Inf, 
             label = paste("Skew:", skew_val, "\nMean:", mean_val, "\nMed:", median_val),
             hjust = 1.1, vjust = 1.1, size = 3, color = "black") +
    theme_minimal() +
    labs(title = var, x = var)
})

# Display histograms
do.call(grid.arrange, c(hist_with_stats, ncol = 3))

# Identify specific outliers for highly skewed variables
find_outlier_values <- function(x, var_name) {
  Q1 <- quantile(x, 0.25)
  Q3 <- quantile(x, 0.75)
  IQR <- Q3 - Q1
  lower_bound <- Q1 - 1.5 * IQR
  upper_bound <- Q3 + 1.5 * IQR
  outlier_indices <- which(x < lower_bound | x > upper_bound)
  
  if(length(outlier_indices) > 0) {
    return(data.frame(
      Variable = var_name,
      Index = outlier_indices,
      Value = x[outlier_indices]
    ))
  } else {
    return(NULL)
  }
}

# Get all outliers
all_outliers <- do.call(rbind, lapply(predictors, function(var) {
  find_outlier_values(Glass[[var]], var)
}))


```

**Highly Skewed (Right/Positive):**

K: Skewness = 6.5 (extremely right-skewed - most values low, few very
high)

Ba: Skewness = 4.2 (extremely right-skewed - mostly zeros)

Fe: Skewness = 2.8 (highly right-skewed)

Al: Skewness = 1.2 (highly right-skewed)

**Moderately Skewed:**

Mg: Slight left/negative skew (median \> mean)

Ca: Moderate right skew

**Approximately Symmetric:**

RI: Very symmetric (mean = median)

Na: Nearly symmetric

Si: Nearly symmetric

***Key Observations:***

K, Ba, and Fe have the most extreme skewness with many zero values Mg is
unique with slight negative skew (higher values more common) RI and Si
are the most normally distributed predictors The presence of zeros in
multiple predictors (Mg, K, Ba, Fe) suggests these elements are absent
in some glass types

(c) **Are there any relevant transformations of one or more predictors
    that might improve the classification model?**

```{r}


vars <- c("K", "Ba", "Fe", "Al", "Ca")

for(var in vars) {
  x <- Glass[[var]]
  x <- x + 0.0001
  bc <- boxcox(lm(x ~ 1), lambda = seq(-2, 2, 0.1), plotit = FALSE)
  lambda <- bc$x[which.max(bc$y)]
  cat(var, "lambda =", round(lambda, 3), "\n")
}
```

-   K (lambda = 0.4): Close to square root (lambda = 0.5) - mild
    transformation

-   Ba (lambda = -0.6): Inverse transformation (1/x\^0.6) - strong
    compression of high values

-   Fe (lambda = -0.3): Mild inverse transformation

-   Al (lambda = 0.5): Perfect square root transformation

-   Ca (lambda = -1.1): Close to inverse (1/x) - very strong
    transformation

**Visualize transformation**

```{r}

# Set specific parameters for this plot
# Visualize why we transform - USING title() which centers by default
par(mfrow = c(2, 2))

for(var in vars) {
  x <- Glass[[var]]
  skew_orig <- round(moments::skewness(x), 2)
  
  hist(x, main = paste(var, "- Original"), 
       col = "steelblue", breaks = 20,
       xlab = var, probability = TRUE)
  lines(density(x), col = "red", lwd = 2)
  
  # Add skewness as subtitle (centered by default)
  title(sub = paste("Skew =", skew_orig), 
        col.sub = "red", cex.sub = 0.9)
}

par(mfrow = c(1, 1))

```


```{r}
par(mfrow = c(2, 2))

for(var in vars) {
  x <- Glass[[var]]
  x_adj <- x + 0.0001  # Handle zeros
  
  # Apply transformation with your lambda
  lambda <- switch(var,
                   "K" = 0.4,
                   "Ba" = -0.6,
                   "Fe" = -0.3,
                   "Al" = 0.5,
                   "Ca" = -1.1)
  
  if(abs(lambda) < 0.01) {
    x_trans <- log(x_adj)
  } else {
    x_trans <- (x_adj^lambda - 1) / lambda
  }
  
  hist(x_trans, main = paste(var, "- Transformed (lambda =", lambda, ")"), 
       col = "coral", breaks = 20,
       xlab = var, probability = TRUE)
  lines(density(x_trans), col = "blue", lwd = 2)
  
  # Add new skewness
  skew_new <- round(moments::skewness(x_trans), 2)
  legend("topright", legend = paste("Skew =", skew_new), 
         bty = "n", text.col = "blue")
}


```


**Box plot before and after transformation**

```{r}
# 2. Compare multiple variables in one figure
par(mfrow = c(2, 2))  # 2 rows, 3 columns

# Variables to transform
vars <- c("K", "Ba", "Fe", "Al", "Ca")
lambdas <- c(0.4, -0.6, -0.3, 0.5, -1.1)  # Your lambda values

for(i in 1:length(vars)) {
  var_name <- vars[i]
  lambda <- lambdas[i]
  
  # Original
  boxplot(Glass[[var_name]] ~ Glass$Type,
          main = paste("Original", var_name),
          xlab = "Type", ylab = var_name,
          col = "steelblue", border = "black",
          cex.axis = 0.7)
  
  # Transformed
  x <- Glass[[var_name]] + 0.0001
  x_trans <- (x^lambda - 1) / lambda
  
  boxplot(x_trans ~ Glass$Type,
          main = paste("Transformed", var_name, "(lambda =", lambda, ")"),
          xlab = "Type", ylab = paste("Transformed", var_name),
          col = "coral", border = "black",
          cex.axis = 0.7)
}

par(mfrow = c(2, 2))
```

**3.2. The soybean data can also be found at the UC Irvine Machine
Learning Repository. Data were collected to predict disease in 683
soybeans. The 35 predictors are mostly categorical and include
information on the environmentalc onditions
(e.g.,temperature,precipitation)and plant conditions (e.g.,left spots,
mold growth). The outcome labels consist of 19 distinct classes.**

**The data can be loaded via:**

```{r}
data(Soybean)
?Soybean
# Check Structure
str(Soybean)

```

(a) **Investigate the frequency distributions for the categorical
    predictors. Are any of the distributions degenerate in the ways
    discussed earlier in this chapter?**

```{r}
# Get categorical predictors (excluding the target variable 'Class')
categorical_vars <- names(Soybean)[names(Soybean) != "Class"]
# Quick check for degenerate distributions
results <- data.frame()

for(var in categorical_vars) {
  # Get frequency table
  tbl <- table(Soybean[[var]])
  
  # Calculate key metrics
  n_levels <- length(tbl)
  most_common <- max(tbl)
  most_common_pct <- round(most_common / sum(tbl) * 100, 2)
  zero_levels <- sum(tbl == 0)
  
  # Store results
  results <- rbind(results, data.frame(
    Variable = var,
    Levels = n_levels,
    Most_Common_Pct = most_common_pct,
    Zero_Freq_Levels = zero_levels,
    row.names = NULL
  ))
}

# Sort by most common percentage (highest first)
results <- results[order(-results$Most_Common_Pct), ]

# Show potential degenerate variables ( >90% in one category or many zero levels)
degenerate <- results[results$Most_Common_Pct > 90 | results$Zero_Freq_Levels > 0, ]
print("Potentially Degenerate Variables:")
print(degenerate)

# Quick view of the worst offenders
print("\nTop 5 most dominant variables:")
print(head(results, 5))

```

(b) **Roughly 18% of the data are missing. Are there particular
    predictors that are more likely to be missing? Is the pattern of
    missing data related to the classes?**

```{r}

# 1. Overall missing data percentage
total_cells <- nrow(Soybean) * ncol(Soybean)
missing_cells <- sum(is.na(Soybean))
missing_pct <- round(missing_cells / total_cells * 100, 2)
cat("Overall missing data:", missing_pct, "%\n")

# 2. Missing data by predictor
missing_by_predictor <- data.frame(
  Variable = names(Soybean),
  Missing_Count = colSums(is.na(Soybean)),
  Missing_Pct = round(colSums(is.na(Soybean)) / nrow(Soybean) * 100, 2)
)

# Sort by missing percentage (highest first)
missing_by_predictor <- missing_by_predictor[order(-missing_by_predictor$Missing_Pct), ]
print("\nMissing data by predictor (top 10):")
print(head(missing_by_predictor, 10))

# 3. Which predictors have the most missing data?
high_missing <- missing_by_predictor[missing_by_predictor$Missing_Pct > 10, ]
print("\nPredictors with >10% missing:")
print(high_missing)

# 4. Missing data by class
missing_by_class <- aggregate(is.na(Soybean[, -which(names(Soybean) == "Class")]), 
                              by = list(Class = Soybean$Class), 
                              FUN = function(x) round(mean(is.na(x)) * 100, 2))

print("\nMissing percentage by class (overall):")
print(data.frame(
  Class = missing_by_class$Class,
  Overall_Missing_Pct = rowMeans(missing_by_class[, -1])
))

# 5. For top missing predictors, check pattern by class
top_missing_vars <- head(missing_by_predictor$Variable[missing_by_predictor$Variable != "Class"], 3)

for(var in top_missing_vars) {
  cat("\n", var, "missing by class:\n")
  missing_by_class_var <- aggregate(is.na(Soybean[[var]]), 
                                    by = list(Class = Soybean$Class), 
                                    FUN = function(x) round(mean(x) * 100, 2))
  names(missing_by_class_var)[2] <- "Missing_Pct"
  print(missing_by_class_var[order(-missing_by_class_var$Missing_Pct), ])
}

# 6. Visualize missing data pattern (simple)

# Prepare data for top variables
top_vars <- head(missing_by_predictor$Variable[missing_by_predictor$Variable != "Class"], 6)
missing_df <- data.frame(
  Class = rep(Soybean$Class, length(top_vars)),
  Variable = rep(top_vars, each = nrow(Soybean)),
  Missing = as.vector(sapply(top_vars, function(v) is.na(Soybean[[v]])))
)

# Calculate missing percentage by class and variable
missing_summary <- aggregate(Missing ~ Class + Variable, 
                             data = missing_df, 
                             FUN = function(x) round(mean(x) * 100, 2))

# Plot
ggplot(missing_summary, aes(x = Class, y = Variable, fill = Missing)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "red") +
  theme_minimal() +
  labs(title = "Missing Data Pattern by Class and Variable",
       fill = "% Missing") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# 7. Quick check: Is missingness related to class?
cat("\nTesting if missingness is related to class (chi-square test):\n")
for(var in top_missing_vars) {
  # Create contingency table: Class vs Missing
  missing_indicator <- is.na(Soybean[[var]])
  tbl <- table(Soybean$Class, missing_indicator)
  
  # Only test if table has sufficient data
  if(min(dim(tbl)) > 1 && all(rowSums(tbl) > 0)) {
    test <- chisq.test(tbl, simulate.p.value = TRUE)
    cat(var, "- p-value:", round(test$p.value, 4), 
        ifelse(test$p.value < 0.05, " (Significant)", " (Not significant)"), "\n")
  } else {
    cat(var, "- insufficient data for test\n")
  }
}
```

#### **Analysis of Missing Data Patterns**

The output reveals a clear pattern of informative missingness (Missing
Not At Random - MNAR). Here's what it means:

**Key Findings:**

**Classes with 100% missing data:**

2-4-d-injury cyst-nematode diaporthe-pod-&-stem-blight herbicide-injury

These 4 disease classes have complete missing data across all
predictors.

This means:

No predictor information is available for these classes They cannot be
used in predictive modeling without imputation They represent distinct
disease types that were measured differently Phytophthora-rot has 77.27%
missing data

Most predictors missing for this class Partial information available All
other 14 classes have 0% missing data

Complete predictor information These will be the well-represented
classes in modeling Statistical Significance: The chi-square tests show
that missingness in predictors (hail, sever, seed.tmt) is significantly
related to class (p \< 0.05). This confirms that missing data is not
random but depends on the disease type.

##### (c) Develop a strategy for handling missing data, either by eliminating predictors or imputation.

##### Handling missing data with Hybrid Strategy

```{r}
# Recommended Hybrid Strategy
hybrid_imputation <- function(data) {
  
  data_clean <- data
  
  # Step 1: Remove classes with >80% missing (can't impute reliably)
  class_missing_pct <- aggregate(is.na(data[, -1]), by = list(Class = data$Class), 
                                  FUN = function(x) mean(is.na(x)) * 100)
  class_avg_missing <- rowMeans(class_missing_pct[, -1])
  classes_to_keep <- class_missing_pct$Class[class_avg_missing < 80]
  
  data_clean <- data_clean[data_clean$Class %in% classes_to_keep, ]
  cat("Classes retained:", length(unique(data_clean$Class)), "\n")
  
  # Step 2: Remove predictors with >40% missing (after class filtering)
  predictor_missing <- colSums(is.na(data_clean[, -1])) / nrow(data_clean) * 100
  predictors_to_keep <- names(predictor_missing[predictor_missing < 40])
  data_clean <- data_clean[, c("Class", predictors_to_keep)]
  
  # Step 3: Class-based mode/median imputation for remaining missing
  for(class_val in unique(data_clean$Class)) {
    class_idx <- which(data_clean$Class == class_val)
    
    for(var in predictors_to_keep) {
      missing_idx <- class_idx[is.na(data_clean[class_idx, var])]
      
      if(length(missing_idx) > 0) {
        # Get non-missing values from same class
        class_vals <- data_clean[class_idx, var]
        non_missing <- class_vals[!is.na(class_vals)]
        
        if(length(non_missing) > 0) {
          if(is.factor(data_clean[[var]])) {
            # Mode for factors
            data_clean[missing_idx, var] <- names(sort(table(non_missing), decreasing = TRUE))[1]
          } else {
            # Median for numeric
            data_clean[missing_idx, var] <- median(non_missing, na.rm = TRUE)
          }
        } else {
          # If class has no non-missing values, use global mode/median
          global_vals <- data_clean[[var]][!is.na(data_clean[[var]])]
          if(is.factor(data_clean[[var]])) {
            data_clean[missing_idx, var] <- names(sort(table(global_vals), decreasing = TRUE))[1]
          } else {
            data_clean[missing_idx, var] <- median(global_vals, na.rm = TRUE)
          }
        }
      }
    }
  }
  
  return(data_clean)
}

# Apply hybrid strategy
Soybean_clean <- hybrid_imputation(Soybean)
cat("\nFinal dataset dimensions:", dim(Soybean_clean))
cat("\nRemaining missing:", sum(is.na(Soybean_clean)))
```

##### Hybrid Imputation Strategy

##### Step 1: Remove Classes with Excessive Missing Data

**Identified classes with \>80% missing data:** "2-4-d-injury",
"cyst-nematode", "diaporthe-pod-&-stem-blight", "herbicide-injury"

**Why?** These classes had nearly all predictors missing (100% for most)

**Result:** Removed these 4 classes completely (they were unusable for
modeling)

**Outcome:** Dataset reduced from 683 to 683 rows (actually these
classes had very few samples, so row count remained similar)

##### Step 2: Remove Problematic Predictors

Calculated missing percentage for each predictor across remaining
classes Removed predictors with \>40% missing (high_missing variables)

**Why?** Imputing \>40% of values would create too much artificial data

**Result:** Removed highly missing predictors from the dataset

**Outcome:** Number of predictors reduced (from original to 36 columns
remaining)

##### Step 3: Class-Based Imputation for Remaining Missing

**Why This Approach Works:**

**Class-Specific Logic:** Different soybean diseases have different
characteristic patterns. Imputing within the same class preserves these
disease-specific characteristics.

**Mode for Categorical:** For variables like "hail" (yes/no) or "sever"
(severity level), using the most common value from the same disease
class maintains the typical pattern.

**Median for Numeric:** For any numeric predictors, median is resistant
to outliers and better represents the "typical" value for that class.
Fallback Strategy: If a class has no non-missing values for a variable,
we use global statistics as backup.

#### Advantages of This Hybrid Approach:

Preserves class characteristics - Imputes based on similar observations
Handles different data types - Appropriate methods for factors vs
numeric Robust to outliers - Uses median instead of mean Multiple
fallback levels - Class-specific first, global second No remaining
missing - Ready for any modeling algorithm Maintains interpretability -
Imputed values are realistic for each disease class

This strategy is particularly effective for the Soybean dataset because
missing patterns are strongly tied to disease class, as confirmed by the
significant chi-square tests we ran earlier.

------------------------------------------------------------------------
