---
title: "Homework3_Data-624"
author: "Mehreen Ali Gillani"
date: "2026-02-19"
output:
  pdf_document: default
  html_document: default
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(fpp3)
library(tidyverse)
library(knitr)
```

#### 5.1) Produce forecasts for the following series using whichever of NAIVE(y), SNAIVE(y) or RW(y \~ drift()) is more appropriate in each case:

Australian Population (global_economy) Bricks (aus_production) NSW Lambs
(aus_livestock) Household wealth (hh_budget). Australian takeaway food
turnover (aus_retail).

##### 1. Australian Population (global_economy)

```{r}

# Filter for Australia
aus_pop <- global_economy %>%
  filter(Country == "Australia") %>%
  select(Year, Population)

# Plot to examine trend
aus_pop %>%
  autoplot(Population) +
  labs(title = "Australian Population", y = "Population")

# Check for trend
aus_pop %>%
  features(Population, features = list(unitroot_ndiffs, unitroot_nsdiffs))
```

**Analysis:** Population data typically shows strong upward trend with
no seasonality (annual data). The most appropriate method would be RW(y
\~ drift()) to capture the growth trend.

```{r}
# Forecast using RW with drift
aus_pop_fit <- aus_pop %>%
  model(RW_with_drift = RW(Population ~ drift()))

aus_pop_fc <- aus_pop_fit %>%
  forecast(h = 10)

aus_pop_fc %>%
  autoplot(aus_pop) +
  labs(title = "Australian Population Forecasts", y = "Population")
```

#### 2. Bricks (aus_production)

```{r}
# Load and prepare bricks data
bricks <- aus_production %>%
  select(Quarter, Bricks) %>%
  filter(!is.na(Bricks)) %>%
  as_tsibble(index = Quarter)

# 1. Time plot to visualize series
bricks %>%
  autoplot(Bricks) +
  labs(title = "Australian Brick Production (1956-2005)",
       y = "Bricks (million units)")

# 2. Seasonal plot to confirm pattern
bricks %>%
  gg_season(Bricks) +
  labs(title = "Seasonal Plot: Quarterly Pattern")

# 3. ACF to verify seasonality
bricks %>%
  ACF(Bricks, lag_max = 16) %>%
  autoplot() +
  labs(title = "ACF: Spikes at lags 4,8,12 confirm seasonality")

# 4. Fit SNAIVE model and forecast
bricks_fit <- bricks %>%
  model(snaive = SNAIVE(Bricks))

# 5. Generate and plot forecasts with intervals
bricks_fc <- bricks_fit %>%
  forecast(h = 8)

bricks_fc %>%
  autoplot(bricks, level = c(80, 95)) +
  labs(title = "Brick Production: SNAIVE Forecast",
       subtitle = "8-quarter forecast with 80% and 95% prediction intervals",
       y = "Bricks (million units)")
```

**Why SNAIVE (Seasonal Naïve) is Appropriate:**

1.  Strong Quarterly Seasonality

2.  How SNAIVE Works

SNAIVE uses the value from the same season in the previous year Formula:
ŷ\_{T+h} = y\_{T+h - m} where m = seasonal period (4 for quarterly) For
example:

**Why Not Other Methods?**

**Why These Method Not Appropriate** NAIVE Would use previous quarter
only (e.g., 2005 Q3 = 2005 Q2 = 435), ignoring seasonal pattern RW with
drift Would impose linear trend, but brick production doesn't have
consistent long-term trend Mean method Would forecast the average (=277)
for all quarters, missing seasonal peaks/troughs

#### NSW Lambs (aus_livestock)

```{r}
# Filter for NSW lambs
nsw_lambs <- aus_livestock %>%
  filter(State == "New South Wales", 
         Animal == "Lambs") %>%
  select(Month, Count)

# Plot to examine pattern
nsw_lambs %>%
  autoplot(Count) +
  labs(title = "NSW Lamb Slaughterings", y = "Count")

# Check for seasonality
nsw_lambs %>%
  gg_season(Count, period = "year") +
  labs(title = "Seasonal Plot: NSW Lamb Slaughterings")

# ACF to confirm seasonality pattern
nsw_lambs %>%
  ACF(Count, lag_max = 36) %>%
  autoplot()

```

**Analysis:** Livestock data shows strong seasonal patterns (breeding
cycles, seasonal demand) with possible trends. SNAIVE(y) would be most
appropriate to capture the monthly seasonal pattern.

```{r}
# Forecast using SNAIVE
nsw_lambs_fit <- nsw_lambs %>%
  model(SNAIVE = SNAIVE(Count))

nsw_lambs_fc <- nsw_lambs_fit %>%
  forecast(h = 24)  # 2 years ahead

nsw_lambs_fc %>%
  autoplot(nsw_lambs) +
  labs(title = "NSW Lamb Slaughterings Forecasts", y = "Count")

```

#### Household wealth (hh_budget).

```{r}
# Household wealth data
hh_wealth <- hh_budget %>%
  select(Year, Wealth)  # Assuming Wealth column exists

# Check if Wealth is the correct variable name
names(hh_budget)

# Plot to examine trend
hh_wealth %>%
  autoplot(Wealth) +
  labs(title = "Household Wealth", y = "Wealth")

# Check for unit roots to confirm trend
hh_wealth %>%
  features(Wealth, unitroot_ndiffs)

```

```{r}
# Forecast using RW with drift
hh_wealth_fit <- hh_wealth %>%
  model(RW_with_drift = RW(Wealth ~ drift()))

hh_wealth_fc <- hh_wealth_fit %>%
  forecast(h = 5)

hh_wealth_fc %>%
  autoplot(hh_wealth) +
  labs(title = "Household Wealth Forecasts", y = "Wealth")

```

**Analysis:** NAIVE would ignore the trend, RW with drift captures the
average annual increase RW with drift uses the last value plus the
average historical change - perfect for trending annual data like
wealth, GDP, or population.

```{r}

# Fit different models
hh_models <- hh_wealth %>%
  model(
    `RW with Drift` = RW(Wealth ~ drift()),
    `Naive` = NAIVE(Wealth),
    `Mean` = MEAN(Wealth)
  )

# Generate forecasts
hh_forecasts <- hh_models %>%
  forecast(h = 5)

# Plot all forecasts together
hh_forecasts %>%
  autoplot(hh_wealth, level = NULL) +
  labs(title = "Household Wealth - Comparison of Different Models",
       subtitle = "5-year forecasts",
       y = "Wealth ($AUD)",
       x = "Year") +
  theme_minimal()

# Individual plots for each model
library(patchwork)

# Create separate plots
p1 <- hh_models %>%
  select(`RW with Drift`) %>%
  forecast(h = 5) %>%
  autoplot(hh_wealth) +
  labs(title = "RW with Drift", y = "Wealth")

p2 <- hh_models %>%
  select(Naive) %>%
  forecast(h = 5) %>%
  autoplot(hh_wealth) +
  labs(title = "Naive", y = "Wealth")

p3 <- hh_models %>%
  select(Mean) %>%
  forecast(h = 5) %>%
  autoplot(hh_wealth) +
  labs(title = "Mean", y = "Wealth")



# Display all plots
p1 + p2 + p3 

```

#### Australian Capital Territory - Takeaway

```{r aus-takeaway}
takeaway <- aus_retail %>%
  filter(State %in% c("Australian Capital Territory")) %>%
  filter(str_detect(Industry, "Takeaway")) %>%  # Partial matching
  select(Month, Turnover)

# Plot to examine pattern
takeaway %>%
  autoplot(Turnover) +
  labs(title = "Australian Takeaway Food Turnover", 
       y = "Turnover ($million)")

# Check for seasonality and trend
takeaway %>%
  gg_season(Turnover, period = "year") +
  labs(title = "Seasonal Plot: Takeaway Food Turnover")

# Decomposition to visualize components
takeaway %>%
  model(classical_decomposition(Turnover, type = "additive")) %>%
  components() %>%
  autoplot()

```

```{r}
# Fit both models
takeaway_fits <- takeaway %>%
  model(
    snaive = SNAIVE(Turnover),
    rw_drift = RW(Turnover ~ drift())
  )

# Generate forecasts (2 years ahead)
takeaway_forecasts <- takeaway_fits %>%
  forecast(h = 24)

# Plot both forecasts together
takeaway_forecasts %>%
  autoplot(takeaway, level = NULL) +
  labs(title = "Australian Takeaway Food Turnover: SNAIVE vs RW with Drift",
       subtitle = "SNAIVE captures seasonality, RW with drift captures trend",
       y = "Turnover ($million)",
       x = "Month") +
  theme_minimal()

# View forecast values
takeaway_forecasts %>%
  as_tibble() %>%
  select(Month, .model, .mean) %>%
  pivot_wider(names_from = .model, values_from = .mean) %>%
  print(n = 24)

```

**SNAIVE (first column): Shows clear seasonal pattern**

Jan: 21.1 (low), Dec: 29.1 (high) Repeats exactly each year (2020
identical to 2019) RW with drift (second column): Just a straight line

Slowly increases: 29.2 → 30.5 over 2 years No seasonal variation at all

**Why SNAIVE looks better:**

Takeaway food HAS strong seasonality (more in summer, less in winter)
The SNAIVE forecasts capture this: Jan/Feb lowest, Dec highest RW with
drift ignores this entirely - same value all year!

### 5.2 Use the Facebook stock price (data set gafa_stock) to do the following:

```{r}
# Load and prepare Facebook stock data
fb_stock <- gafa_stock %>%
  filter(Symbol == "FB") %>%
  select(Date, Close) %>%
  as_tsibble(index = Date)

# Check the data
glimpse(fb_stock)
summary(fb_stock)
# check for interval
interval(fb_stock)
```

#### a. Time Plot of the Series

```{r}
# Time plot of Facebook stock price
p1 <- fb_stock %>%
  autoplot(Close) +
  labs(title = "Facebook Stock Price (2015-2018)",
       subtitle = "Daily closing price",
       y = "Closing Price ($USD)",
       x = "Date") +
  theme_minimal()

print(p1)
```

#### b. Forecasts Using Drift Method

```{r}
# Calculate drift
first_close <- head(fb_stock$Close, 1)
last_close <- tail(fb_stock$Close, 1)
n_days <- nrow(fb_stock)
drift <- (last_close - first_close) / n_days

# Generate future trading days
last_date <- max(fb_stock$Date)
future_dates <- seq(last_date + 1, by = "day", length.out = 60)
future_dates <- future_dates[!weekdays(future_dates) %in% c("Saturday", "Sunday")]
future_dates <- future_dates[1:30]

# Create forecast
fb_fc <- tibble(
  Date = future_dates,
  Drift = last_close + drift * (1:30),
  Naive = last_close,
  Mean = mean(fb_stock$Close)
)

# Plot
fb_stock %>%
  filter(Date >= as.Date("2018-01-01")) %>%
  ggplot(aes(x = Date, y = Close)) +
  geom_line() +
  geom_line(data = fb_fc, aes(x = Date, y = Drift, color = "Drift"), linetype = "dashed") +
  geom_line(data = fb_fc, aes(x = Date, y = Naive, color = "Naive"), linetype = "dashed") +
  geom_line(data = fb_fc, aes(x = Date, y = Mean, color = "Mean"), linetype = "dashed") +
  labs(title = "Facebook Stock Price Forecasts",
       y = "Closing Price ($USD)",
       color = "Method") +
  theme_minimal()

# Show forecasts
print("30-day forecasts:")
print(fb_fc)
```

#### c. Show that the forecasts are identical to extending the line drawn between the first and last observations.

```{r}
# Get first and last observations
first_point <- fb_stock %>% slice_head(n = 1)
last_point <- fb_stock %>% slice_tail(n = 1)

first_date <- first_point$Date
first_price <- first_point$Close
last_date <- last_point$Date
last_price <- last_point$Close
n <- nrow(fb_stock)  # Number of observations

print(paste("First observation:", first_date, "- $", round(first_price, 2)))
print(paste("Last observation:", last_date, "- $", round(last_price, 2)))
print(paste("Total observations:", n))

# Generate future trading days (next 30 trading days)
last_date <- max(fb_stock$Date)
all_dates <- seq(last_date + 1, by = "day", length.out = 60)
trading_days <- all_dates[!weekdays(all_dates) %in% c("Saturday", "Sunday")]
future_dates <- trading_days[1:30]
h <- length(future_dates)  # 30

# Calculate slope between first and last points
# Number of steps between first and last = n - 1
slope <- (last_price - first_price) / (n - 1)

# Create the line data (historical period + forecast period)
line_dates <- c(fb_stock$Date, future_dates)
line_values <- first_price + slope * (0:(length(line_dates) - 1))

# Now create the tibble with correct lengths
line_data <- tibble(
  Date = line_dates,
  Price = line_values,
  Type = c(rep("Historical Line", length(fb_stock$Date)), 
           rep("Extended Line", length(future_dates)))
)

# Check the lengths
print(paste("Length of line_dates:", length(line_dates)))
print(paste("Length of line_values:", length(line_values)))
print(paste("Length of Type:", nrow(line_data)))
print(paste("All match:", length(line_dates) == length(line_values) && 
                          length(line_values) == nrow(line_data)))

# Drift forecast (for comparison)
drift_per_day <- (last_price - first_price) / n
drift_forecast <- tibble(
  Date = future_dates,
  Price = last_price + drift_per_day * (1:h),
  Type = "Drift Forecast"
)

# Plot to show they're on the same line
p <- ggplot() +
  # Original data
  geom_line(data = fb_stock, aes(x = Date, y = Close, color = "Actual Price"), 
            size = 0.8) +
  # Line from first to last (extended)
  geom_line(data = line_data, aes(x = Date, y = Price, color = "Line: First to Last"), 
            linetype = "dashed", size = 1) +
  # Drift forecast (should overlap exactly)
  geom_point(data = drift_forecast, aes(x = Date, y = Price, color = "Drift Forecast"), 
             size = 2, shape = 1) +
  # Mark first and last points
  geom_point(data = first_point, aes(x = Date, y = Close), 
             color = "blue", size = 4) +
  geom_point(data = last_point, aes(x = Date, y = Close), 
             color = "red", size = 4) +
  scale_color_manual(values = c("Actual Price" = "black", 
                                 "Line: First to Last" = "red",
                                 "Drift Forecast" = "blue")) +
  labs(title = "Drift Forecast = Extended Line from First to Last Point",
       subtitle = paste("Line from", format(first_date, "%Y-%m-%d"), "($", round(first_price, 2), 
                       ") to", format(last_date, "%Y-%m-%d"), "($", round(last_price, 2), ")"),
       y = "Price ($USD)",
       x = "Date",
       color = "Legend") +
  theme_minimal() +
  theme(legend.position = "bottom")

print(p)
# Zoom in on the last part to see clearly
# Zoom in on the last part to see clearly
p_zoom <- ggplot() +
  # Last 6 months of actual data
  geom_line(data = fb_stock %>% filter(Date >= as.Date("2018-07-01")), 
            aes(x = Date, y = Close, color = "Actual"), size = 1) +
  
  # Complete line from first point (using line_data, NOT fb_stock)
  geom_line(data = line_data %>% filter(Date >= as.Date("2018-07-01")), 
            aes(x = Date, y = Price, color = "Line from First Point"), 
            linetype = "dashed", size = 1) +
  
  # Drift forecast
  geom_point(data = drift_forecast, 
             aes(x = Date, y = Price, color = "Drift Forecast"), 
             size = 4) +
  
  # Last point
  geom_point(data = last_point, aes(x = Date, y = Close), 
             color = "red", size = 5) +
  
  scale_color_manual(values = c("Actual" = "black", 
                                 "Line from First Point" = "red",
                                 "Drift Forecast" = "blue")) +
  
  labs(title = "Zoomed: Drift Forecast Continues the Line",
       subtitle = "Blue dots should lie exactly on red dashed line",
       y = "Price ($USD)") +
  
  theme_minimal()

print(p_zoom)
```

#### d. Try using some of the other benchmark functions to forecast the same data set. Which do you think is best? Why?

```{r}

# 2. Make it regular (daily)
fb_regular <- data.frame(
  Date = seq(min(fb_stock$Date), max(fb_stock$Date), by = "day")
) %>%
  left_join(fb_stock, by = "Date") %>%
  arrange(Date) %>%
  mutate(
    Close = zoo::na.locf(Close, na.rm = FALSE),
    Close = zoo::na.locf(Close, fromLast = TRUE, na.rm = FALSE)
  ) %>%
  as_tsibble(index = Date)

print(paste("Regular data created:", nrow(fb_regular), "days"))

# 3. Fit benchmark models
fb_models <- fb_regular %>%
  model(
    naive = NAIVE(Close),
    rw = RW(Close),
    drift = RW(Close ~ drift()),
    mean = MEAN(Close)
  )

# 4. Generate forecasts
fb_fc <- fb_models %>%
  forecast(h = 30)

# 5. Plot all forecasts
p_all <- fb_fc %>%
  autoplot(fb_regular %>% filter(Date >= as.Date("2018-01-01")), level = NULL) +
  labs(title = "Facebook Stock Price - Benchmark Method Comparison",
       subtitle = "30-day forecasts (regularized daily data)",
       y = "Closing Price ($USD)") +
  theme_minimal()

print(p_all)

# 6. Calculate accuracy with cross-validation
fb_cv <- fb_regular %>%
  stretch_tsibble(.init = 1000, .step = 100) %>%
  model(
    naive = NAIVE(Close),
    drift = RW(Close ~ drift()),
    mean = MEAN(Close)
  )

fb_cv_fc <- fb_cv %>%
  forecast(h = 30)

accuracy_table <- fb_cv_fc %>%
  accuracy(fb_regular) %>%
  select(.model, RMSE, MAE, MAPE) %>%
  arrange(RMSE)

print("Model Accuracy Comparison:")
print(accuracy_table)

# 7. Test random walk hypothesis
fb_returns <- fb_regular %>%
  mutate(return = (Close - lag(Close)) / lag(Close) * 100) %>%
  drop_na()

# Ljung-Box test
lb_test <- Box.test(fb_returns$return, lag = 10, type = "Ljung-Box")
print(paste("Ljung-Box p-value:", round(lb_test$p.value, 4)))

# 8. Determine best method
best_method <- accuracy_table %>%
  slice_min(RMSE) %>%
  pull(.model)

print(paste("Best method based on RMSE:", best_method))

# 9. Generate best forecast
best_forecast <- fb_models %>%
  select(!!best_method) %>%
  forecast(h = 30)

# 10. Plot best forecast with intervals
best_forecast %>%
  autoplot(fb_regular %>% filter(Date >= as.Date("2018-06-01")), level = c(80, 95)) +
  labs(title = paste("Facebook Stock Price - Best Method:", best_method),
       subtitle = paste("30-day forecast - Ljung-Box p-value:", round(lb_test$p.value, 4)),
       y = "Closing Price ($USD)") +
  theme_minimal()

# 11. Show forecast values
forecast_values <- best_forecast %>%
  as_tibble() %>%
  select(Date, .mean) %>%
  mutate(Date = as.Date(Date))

print("30-day forecast:")
print(forecast_values)

```

***The results show that Drift method performs best for Facebook stock
with the lowest RMSE (7.08 vs 7.32 for Naive).***

##### Further Analysis

```{r}
# Calculate the overall trend
first_price <- fb_regular$Close[1]
last_price <- tail(fb_regular$Close, 1)
n_days <- nrow(fb_regular)
drift_per_day <- (last_price - first_price) / n_days

print(paste("First price (2014): $", round(first_price, 2)))
print(paste("Last price (2018): $", round(last_price, 2)))
print(paste("Total increase: $", round(last_price - first_price, 2)))
print(paste("Drift per day: $", round(drift_per_day, 4)))
print(paste("Drift per year: $", round(drift_per_day * 252, 2)))  # ~252 trading days

# Visualize the trend
fb_regular %>%
  ggplot(aes(x = Date, y = Close)) +
  geom_line(color = "black", alpha = 0.5) +
  geom_abline(intercept = first_price, slope = drift_per_day, 
              color = "red", size = 1, linetype = "dashed") +
  labs(title = "Facebook Stock with Linear Trend Line",
       subtitle = paste("Drift = $", round(drift_per_day, 4), "per day"),
       y = "Closing Price ($USD)")
```

```{r}
# Generate both forecasts
last_close <- tail(fb_regular$Close, 1)
forecast_dates <- seq(max(fb_regular$Date) + 1, by = "day", length.out = 30)

comparison <- tibble(
  Date = forecast_dates,
  Drift = last_close + drift_per_day * (1:30),
  Naive = last_close,
  Actual = NA  # We don't have actuals for future
) %>%
  pivot_longer(-Date, names_to = "Method", values_to = "Forecast")

# Plot comparison
fb_regular %>%
  filter(Date >= as.Date("2018-06-01")) %>%
  ggplot(aes(x = Date, y = Close)) +
  geom_line(color = "black", size = 1) +
  geom_line(data = comparison, aes(x = Date, y = Forecast, color = Method, linetype = Method),
            size = 1) +
  scale_color_manual(values = c("Drift" = "red", "Naive" = "blue")) +
  labs(title = "Drift vs Naive Forecasts",
       subtitle = paste("Drift method expects continued upward trend of $", round(drift_per_day, 4), "per day"),
       y = "Closing Price ($USD)") +
  theme_minimal()
```

```{r}
# Create detailed comparison
forecast_details <- tibble(
  Day = 1:30,
  Date = forecast_dates,
  `Drift Forecast` = round(last_close + drift_per_day * (1:30), 2),
  `Naive Forecast` = round(last_close, 2),
  `Difference` = round(drift_per_day * (1:30), 2),
  `Cumulative Drift` = round(cumsum(rep(drift_per_day, 30)), 2)
)

print("30-Day Forecast Comparison:")
print(forecast_details)

# Plot the difference
forecast_details %>%
  ggplot(aes(x = Day, y = Difference)) +
  geom_col(fill = "darkgreen", alpha = 0.7) +
  labs(title = "Drift Method Premium over Naive",
       subtitle = paste("Drift adds $", round(drift_per_day, 4), "per day"),
       x = "Days Ahead", y = "Additional $ Forecasted") +
  theme_minimal()

```

#### **Analysis of Facebook Stock Results**

***Key Statistics:***

First price (2014): \$54.71 Last price (2018): \$131.09

Total increase: \$76.38 over \~5 years

Drift per day: \$0.0419 (about 4 cents per day)

Drift per year: \$10.55 (annualized trend

The drift forecast shows a gradual increase: Day 1: \$131.09 + \$0.04 =
\$131.13

Day 10: \$131.09 + \$0.42 = \$131.51

Day 30: \$131.09 + \$1.26 = \$132.35

### 5.3 Apply a seasonal naïve method to the quarterly Australian beer production data from 1992. Check if the residuals look like white noise, and plot the forecasts. The following code will help.

***What do you conclude?***

```{r}
# Extract data of interest
recent_production <- aus_production |>
  filter(year(Quarter) >= 1992)
fit <- recent_production |> model(SNAIVE(Beer))

# Generate the residual plots
residual_plots <- fit |> gg_tsresiduals()
print(residual_plots)
```

```{r}
# Extract data of interest
recent_production <- aus_production |>
  filter(year(Quarter) >= 1992)
# Define and estimate a model
fit <- recent_production |> model(SNAIVE(Beer))
# Look at the residuals
fit |> gg_tsresiduals()
# Look a some forecasts
fit |> forecast() |> autoplot(recent_production)
```

***RESIDUAL ANALYSIS FOR SEASONAL NAÏVE MODEL (BEER PRODUCTION)***

1.  TIME PLOT OF RESIDUALS: The residual plot reveals clear seasonal
    patterns, with values alternating above and below zero in a
    systematic way. This cyclical behavior indicates that the model has
    not fully captured the seasonal structure in the data.

2.  ACF PLOT: The autocorrelation function shows 4 lags with bars
    extending beyond the blue dashed confidence bands, with an
    additional 2 bars (lags 1 and 3) touching the boundary. This is
    substantially more than the 0-1 significant lags expected by random
    chance, confirming significant autocorrelation in the residuals.

3.  HISTOGRAM: The residual distribution ranges from approximately -25
    to +25, with a few values extending below -25 suggesting mild left
    skewness. However, the deviation from normality is not severe.

**CONCLUSION:**

-   The residuals are NOT white noise. Despite being roughly normally
    distributed, they exhibit significant autocorrelation (confirmed by
    both the ACF plot).

-   The seasonal patterns visible in the time plot and the multiple
    significant ACF bars indicate that the seasonal naïve model has
    failed to capture all the structure in the beer production data.

-   A more sophisticated model (e.g., ETS or ARIMA with trend
    components) would likely produce better forecasts with white noise
    residuals.

```{r}
# Get the residuals
residuals_df <- fit |> augment()

# Plot ACF of residuals
residuals_df %>%
  ACF(.resid) %>%
  autoplot() +
  labs(title = "ACF of Residuals",
       subtitle = "Should show no significant autocorrelation for white noise")

# Ljung-Box test for white noise
lb_test <- residuals_df %>%
  features(.resid, ljung_box, lag = 8)

print("Ljung-Box test results:")
print(lb_test)

# Interpret the p-value
p_value <- lb_test$lb_pvalue
cat("\nLjung-Box p-value:", round(p_value, 4))
if(p_value > 0.05) {
  cat("\n✓ p-value > 0.05: Residuals appear to be white noise (no significant autocorrelation)")
} else {
  cat("\n✗ p-value < 0.05: Residuals show significant autocorrelation (not white noise)")
}

# Check distribution of residuals
residuals_df %>%
  ggplot(aes(x = .resid)) +
  geom_histogram(aes(y = after_stat(density)), bins = 20, fill = "steelblue", alpha = 0.7) +
  geom_density(color = "red", size = 1) +
  stat_function(fun = dnorm, args = list(mean = 0, sd = sd(residuals_df$.resid)), 
                color = "darkgreen", linetype = "dashed", size = 1) +
  labs(title = "Distribution of Residuals",
       subtitle = "Should be approximately normal for white noise",
       x = "Residuals", y = "Density") +
  theme_minimal()

```

***Key Findings:*** The Ljung-Box test returns a p-value of 0.0001,
which is well below the 0.05 significance threshold.

This indicates significant autocorrelation remains in the residuals,
meaning they are NOT white noise. Therefore, the seasonal naïve model is
inadequate for this beer production data.

### 5.4 Repeat the previous exercise using the Australian Exports series from global_economy and the Bricks series from aus_production. Use whichever of NAIVE() or SNAIVE() is more appropriate in each case.

#### Australian Exports series from global_economy

```{r 5.4}
# Get Australian exports data
aus_exports <- global_economy %>%
  filter(Country == "Australia") %>%
  select(Year, Exports)

# Plot the data
aus_exports %>%
  autoplot(Exports) +
  labs(title = "Australian Exports (% of GDP)", y = "Exports")

# Check for seasonality (annual data - no seasonality)
# Use NAIVE or RW with drift?
aus_exports %>%
  autoplot(Exports)  # Shows trend, no seasonality

# Fit appropriate model (NAIVE with drift for trending annual data)
fit_exports <- aus_exports %>%
  model(Drift = RW(Exports ~ drift()))

# Check residuals
fit_exports %>% gg_tsresiduals()

# Ljung-Box test
fit_exports %>% augment() %>% features(.resid, ljung_box, lag = 10)

# Generate forecasts
fit_exports %>% forecast(h = 5) %>% autoplot(aus_exports)


```

***Key Findings:***

-   The residual time plot shows some seasonal patterns.

-   ACF has only one lag slightly outside the confidence bands.

-   The residuals are normally distributed (range -2 to 2) Ljung-Box
    test (p = 0.09 \> 0.05) confirms no significant autocorrelation,
    indicating the drift model is adequate for Australian exports.

#### Bricks data

```{r}

# Get bricks data (quarterly)
bricks <- aus_production %>%
  select(Quarter, Bricks) %>%
  filter(!is.na(Bricks))

# Plot the data
bricks %>%
  autoplot(Bricks) +
  labs(title = "Australian Brick Production", y = "Million units")

# Check for seasonality
bricks %>% gg_season(Bricks)  # Clear quarterly pattern

# Fit SNAIVE (appropriate for quarterly seasonal data)
fit_bricks <- bricks %>%
  model(SNAIVE = SNAIVE(Bricks))

# Check residuals
fit_bricks %>% gg_tsresiduals()

# Ljung-Box test
fit_bricks %>% augment() %>% features(.resid, ljung_box, lag = 8)

# Generate forecasts
fit_bricks %>% forecast(h = 8) %>% autoplot(bricks)

```

#### **Key Findings:** 

-   The residuals show clear problems: the Ljung-Box test (p = 0)
    confirms significant autocorrelation, with many ACF lags outside the
    confidence bands.

-   The residual histogram ranges from -100 to +100 with extreme
    outliers near -200, indicating the SNAIVE model is inadequate for
    brick production data.

**Conclusions:**

1.  NAIVE with drift is appropriate for Australian exports as it shows
    annual data with trend but no seasonality. The Ljung-Box test (p =
    0.09) confirms residuals are white noise, indicating the model
    adequately captures the patterns.

2.  SNAIVE is appropriate for bricks data due to strong quarterly
    seasonality. However, the residuals show significant autocorrelation
    (p = 0) and outliers, suggesting the model misses some patterns
    despite capturing the seasonal component.

### 5.7 For your retail time series (from Exercise 7 in Section 2.10):

#### a. Create a training dataset consisting of observations before 2011 using

```{r 5.7.a}
set.seed(12345)
myseries <- aus_retail |>
  filter(`Series ID` == sample(aus_retail$`Series ID`,1))
myseries_train <- myseries |>
  filter(year(Month) < 2011)
```

#### b. Check that your data have been split appropriately by producing the following plot.

```{r 5.7.b}
autoplot(myseries, Turnover) +
  autolayer(myseries_train, Turnover, colour = "red")
```

#### c. Fit a seasonal naïve model using SNAIVE() applied to your training data (myseries_train).

```{r}
fit <- myseries_train |>
  model(snaive = SNAIVE(Turnover))
```

#### d. Check the residuals.

```{r}
fit |> gg_tsresiduals()
```

#### Do the residuals appear to be uncorrelated and normally distributed?

***No, the residuals do NOT appear to be uncorrelated and normally
distributed.***

-   The ACF plot shows many lines outside the dashed confidence bands,
    indicating significant autocorrelation (correlated residuals).

-   While the histogram is roughly normal between -20 and +20, the
    presence of outliers at 40 and 50 suggests departures from
    normality.

-   The residual time plot also shows cyclical patterns, confirming the
    model has failed to capture all structure in the data.

#### e. Produce forecasts for the test data

```{r}
fc <- fit |>
  forecast(new_data = anti_join(myseries, myseries_train))
fc |> autoplot(myseries)
```

#### f. Compare the accuracy of your forecasts against the actual values.

```{r}
# Simple but clean table
accuracy_table <- bind_rows(
  Training = fit |> accuracy(),
  Test = fc |> accuracy(myseries),
  .id = "Dataset"
) %>%
  select(Dataset, RMSE, MAE, MAPE, MASE) %>%
  mutate(across(where(is.numeric), ~round(., 2)))

print(accuracy_table)

# Or with knitr for better formatting

kable(accuracy_table, caption = "Training vs Test Accuracy", digits = 2)
```

#### g. How sensitive are the accuracy measures to the amount of training data used?

The accuracy measures show relatively low sensitivity to the training
data amount in this case.

-   The test set performance (RMSE = 11.76, MAE = 9.25) is very close to
    the training set performance (RMSE = 11.85, MAE = 8.93), indicating
    the model generalizes well.

-   Interestingly, the MAPE improved from 9.64% on training to 5.55% on
    test, suggesting the test period may have been more stable or
    predictable.

-   The MASE of 1.04 on test data confirms the model performs nearly as
    well as a naive method. Overall, the model appears robust to the
    training data size, with no evidence of severe overfitting or
    underfitting.

    \_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_
