---
title: "Homework 8 Non-Linear models - Data 624"
author: "Mehreen Ali Gillani"
date: "2026-04-20"
output:
  pdf_document: default
  html_document:
    code_folding: hide
    theme: default
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      fig.width = 8, 
                      fig.height = 6, 
                      fig.align = "center",
                      warning = FALSE,
                      message = FALSE)
# Load required libraries
library(tidyverse)
library(ggplot2)
library(gridExtra)
library(mlbench)
library(caret)
library(earth)
library(randomForest)
library(kernlab)
```

*Do problems 7.2 and 7.5 in Kuhn and Johnson. There are only two but they have many parts. Please submit both a link to your Rpubs and the .rmd file.*

#### 7.2. Friedman (1991) introduced several benchmark data sets create by simulation. One of these simulations used the following nonlinear equation to create data:

y = 10sin(πx1x2)+20(x3−0.5)2 +10x4 +5x5 +N(0,σ2)

where the x values are random variables uniformly distributed between [0, 1] (there are also 5 other non-informative variables also created in
the simula tion). The package mlbench contains a function called mlbench.

postResample(pred = knnPred, obs = testData$y) RMSE Rsquared 3.2286834
0.6871735 Which models appear to give the best performance? Does MARS
select the informative predictors (those named X1–X5)?

```{r}

# Set seed for reproducibility
set.seed(200)

# Generate training data
trainingData <- mlbench.friedman1(200, sd = 1)
trainingData$x <- data.frame(trainingData$x)

# Generate test data
testData <- mlbench.friedman1(5000, sd = 1)
testData$x <- data.frame(testData$x)

# Define train control (use repeated CV for stability)
ctrl <- trainControl(method = "repeatedcv", number = 10, repeats = 3)

# 1. kNN (already shown, but rerun for consistency)
knnModel <- train(x = trainingData$x, y = trainingData$y,
                  method = "knn",
                  preProc = c("center", "scale"),
                  tuneLength = 10,
                  trControl = ctrl)

# 2. MARS
marsModel <- train(x = trainingData$x, y = trainingData$y,
                   method = "earth",
                   tuneLength = 10,
                   trControl = ctrl)

# 3. Random Forest
rfModel <- train(x = trainingData$x, y = trainingData$y,
                 method = "rf",
                 tuneLength = 5,
                 trControl = ctrl)

# 4. SVM Radial
svmModel <- train(x = trainingData$x, y = trainingData$y,
                  method = "svmRadial",
                  preProc = c("center", "scale"),
                  tuneLength = 10,
                  trControl = ctrl)

# 5. Neural Network
nnetModel <- train(x = trainingData$x, y = trainingData$y,
                   method = "nnet",
                   preProc = c("center", "scale"),
                   tuneLength = 5,
                   trace = FALSE,
                   trControl = ctrl)

# Collect models
models <- list(kNN = knnModel, MARS = marsModel, RF = rfModel,
               SVM = svmModel, NNET = nnetModel)

# Compare resampling results
resamp_results <- resamples(models)
summary(resamp_results)

# Test set evaluation
test_results <- data.frame(Model = character(),
                           RMSE = numeric(),
                           Rsquared = numeric())

for (name in names(models)) {
  pred <- predict(models[[name]], newdata = testData$x)
  perf <- postResample(pred = pred, obs = testData$y)
  test_results <- rbind(test_results, data.frame(Model = name,
                                                 RMSE = perf[1],
                                                 Rsquared = perf[2]))
}

print(test_results)
```

**Best model: MARS**

- RMSE = 1.81 (much lower than others)

- R² = 0.87 (explains 87% of variance)

**Why MARS Wins?**

- MARS has the lowest RMSE in both resampling and test.

- It also has the highest R² (0.892 average in resampling, 0.868 on
  test).

- MARS handles interactions naturally

*Check variable importance in the MARS model:*

```{r}
varImp(marsModel)
# or
summary(marsModel$finalModel)
```

**1. Does MARS select the informative predictors (X1–X5)?**

*Yes — strongly.*

**From the Importance section:**

Importance: X1, X4, X2, X5, X3, all have nonzero importance (X1 = 100,
X4 = 82, X2 = 63, X5 = 38, X3 = 25.8) X6, X7-unused, X8-unused,
X9-unused, X10-unused

MARS successfully identified all 5 informative predictors and correctly
ignored (or nearly ignored) the 5 non-informative ones.

#### 7.5 Exercise 6.3 describes data for a chemical manufacturing process. Use the same data imputation, data splitting, and pre-processing steps as before and train several nonlinear regression models.

(a) **Which nonlinear regression model gives the optimal resampling and
    test set performance?**

```{r}
# ============================================================
# Exercise 7.5 - Nonlinear Regression Models
# Chemical Manufacturing Process Dataset
# Based on imputation, splitting, and pre-processing from 6.3
# ============================================================



# ============================================================
# Step 1: Load Data (as in 6.3)
# ============================================================
library(AppliedPredictiveModeling)
data(ChemicalManufacturingProcess)


# The data: processPredictors (176x57), yield (176x1)
# 57 predictors: 12 biological + 45 manufacturing process

# Combine into one data frame for easier handling
chem_data <- ChemicalManufacturingProcess
# Note: The dataset already has column name "Yield"
head(chem_data, 2)


cat("Dataset dimensions:", dim(chem_data), "\n")
cat("Missing values:", sum(is.na(chem_data)), "\n")
```

```{r impute}

# ============================================================
# Step 2: Impute Missing Values (as in 6.3)
# ============================================================
# Using k-nearest neighbors imputation (common approach from 6.3)

# KNN Imputation (most common in solutions)
set.seed(123)
impute_preproc <- preProcess(chem_data, 
                              method = c("knnImpute"), 
                              k = 5)

chem_imputed <- predict(impute_preproc, chem_data)

# Verify no missing values remain
cat("Missing values after imputation:", sum(is.na(chem_imputed)), "\n")

```

```{r train/test split}
# ============================================================
# Step 3: Split Data into Training and Test Sets
# ============================================================
set.seed(517)  # Using seed from exercise 6.3 solutions

train_index <- createDataPartition(chem_imputed$Yield, 
                                    p = 0.8, 
                                    list = FALSE)

train_data <- chem_imputed[train_index, ]
test_data <- chem_imputed[-train_index, ]

cat("Training set size:", nrow(train_data), "\n")
cat("Test set size:", nrow(test_data), "\n")

# Separate predictors and response
train_x <- train_data %>% select(-Yield)
train_y <- train_data$Yield
test_x <- test_data %>% select(-Yield)
test_y <- test_data$Yield
```

```{r}
# ============================================================
# Step 4: Define Training Control
# ============================================================
# Using repeated cross-validation for more stable estimates
ctrl <- trainControl(method = "repeatedcv", 
                     number = 10, 
                     repeats = 3,
                     verboseIter = FALSE)

# Pre-processing options (center + scale for all models)
preProc_options <- c("center", "scale")

```

```{r Train_Multiple_Nonlinear_Regression_Models}
# ============================================================
# Step 5: Train Multiple Nonlinear Regression Models
# ============================================================

# ---------- 5.1 k-Nearest Neighbors (KNN) ----------
set.seed(456)
knn_model <- train(x = train_x,
                   y = train_y,
                   method = "knn",
                   preProc = preProc_options,
                   tuneLength = 10,
                   trControl = ctrl)

cat("\n=== KNN Best Tuning Parameters ===\n")
print(knn_model$bestTune)

# ---------- 5.2 Neural Network (NNET) ----------
set.seed(456)
nnet_grid <- expand.grid(decay = c(0, 0.01, 0.1),
                          size = c(1, 3, 5, 7))

nnet_model <- train(x = train_x,
                    y = train_y,
                    method = "nnet",
                    preProc = preProc_options,
                    tuneGrid = nnet_grid,
                    trControl = ctrl,
                    linout = TRUE,
                    trace = FALSE,
                    maxit = 500)

cat("\n=== Neural Network Best Tuning Parameters ===\n")
print(nnet_model$bestTune)

# ---------- 5.3 Random Forest (RF) ----------
set.seed(456)
rf_model <- train(x = train_x,
                  y = train_y,
                  method = "rf",
                  preProc = preProc_options,
                  tuneLength = 10,
                  trControl = ctrl,
                  importance = TRUE)

cat("\n=== Random Forest Best Tuning Parameters ===\n")
print(rf_model$bestTune)

# ---------- 5.4 MARS (Earth) ----------
set.seed(456)
mars_grid <- expand.grid(degree = 1:2,
                          nprune = 2:30)

mars_model <- train(x = train_x,
                    y = train_y,
                    method = "earth",
                    preProc = preProc_options,
                    tuneGrid = mars_grid,
                    trControl = ctrl)

cat("\n=== MARS Best Tuning Parameters ===\n")
print(mars_model$bestTune)

# ---------- 5.5 Support Vector Machine (SVM) ----------
set.seed(456)
svm_model <- train(x = train_x,
                   y = train_y,
                   method = "svmRadial",
                   preProc = preProc_options,
                   tuneLength = 15,
                   trControl = ctrl)

cat("\n=== SVM Best Tuning Parameters ===\n")
print(svm_model$bestTune)

```

```{r Evaluate_models}
# ============================================================
# Step 6: Evaluate Models on Training (Resampling) and Test Sets
# ============================================================

# Function to get training performance from train object
get_training_performance <- function(model) {
  results <- model$results
  best_row <- which.min(results$RMSE)
  data.frame(RMSE = results$RMSE[best_row],
             Rsquared = results$Rsquared[best_row])
}

# Function to evaluate on test set
evaluate_test <- function(model, test_x, test_y) {
  predictions <- predict(model, newdata = test_x)
  metrics <- postResample(pred = predictions, obs = test_y)
  data.frame(RMSE = metrics[1], Rsquared = metrics[2])
}

# Collect results
models_list <- list(KNN = knn_model,
                    NeuralNet = nnet_model,
                    RandomForest = rf_model,
                    MARS = mars_model,
                    SVM = svm_model)

# Store results
training_results <- data.frame()
test_results <- data.frame()

for (name in names(models_list)) {
  train_perf <- get_training_performance(models_list[[name]])
  train_perf$Model <- name
  training_results <- rbind(training_results, train_perf)
  
  test_perf <- evaluate_test(models_list[[name]], test_x, test_y)
  test_perf$Model <- name
  test_results <- rbind(test_results, test_perf)
}

# Reorder columns
training_results <- training_results[, c("Model", "RMSE", "Rsquared")]
test_results <- test_results[, c("Model", "RMSE", "Rsquared")]

# Sort by test RMSE (lower is better)
test_results <- test_results[order(test_results$RMSE), ]

cat("\n==================================================\n")
cat("=== TRAINING (Resampling) Performance ===\n")
cat("==================================================\n")
print(training_results[order(training_results$RMSE), ])

cat("\n==================================================\n")
cat("=== TEST SET Performance ===\n")
cat("==================================================\n")
print(test_results)

# ============================================================
# Step 7: Answer to Question (a)
# ============================================================

cat("\n==================================================\n")
cat("=== ANSWER TO QUESTION 7.5(a) ===\n")
cat("==================================================\n")

best_train_model <- training_results[which.min(training_results$RMSE), "Model"]
best_test_model <- test_results[which.min(test_results$RMSE), "Model"]

cat("Optimal resampling (training) performance:", best_train_model, "\n")
cat("Optimal test set performance:", best_test_model, "\n")

```

**Optimal Performance:**

**1. SVM performed best on test data**

- Test RMSE = 0.515 (lowest by far)
- Test R² = 0.750 (highest, tied with RF)
- Training RMSE = 0.632 (good, not overfitting)
- Best overall model for generalization

**2. Random Forest was most consistent**

- Test RMSE = 0.542 (second best)
- Test R² = 0.753 (highest)
- Training RMSE = 0.627 (best in resampling)
- Very stable between training and test

**3. MARS, KNN, and Neural Network underperformed**

MARS: Test RMSE 0.615, R² 0.648 KNN: Test RMSE 0.658, R² 0.646
NeuralNet: Test RMSE 0.742, R² 0.595 (worst by far)

```{r Visual}
# ============================================================
#  Visual Comparison
# ============================================================

# Create comparison plot
comparison_df <- rbind(
  data.frame(Model = test_results$Model, Metric = "RMSE", Value = test_results$RMSE, Type = "Test"),
  data.frame(Model = training_results$Model, Metric = "RMSE", Value = training_results$RMSE, Type = "Training")
)

library(ggplot2)
ggplot(comparison_df, aes(x = reorder(Model, -Value), y = Value, fill = Type)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Model Performance Comparison - Chemical Manufacturing Process",
       x = "Model", y = "RMSE (lower is better)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

(b) **Which predictors are most important in the optimal nonlinear
    regression model? Do either the biological or process variables
    dominate the list? How do the top ten important predictors compare
    to the top ten predictors from the optimal linear model?**

```{r}
# ============================================================
# Top 5 Predictors for All Nonlinear Models
# ============================================================

library(tidyverse)
library(caret)

# Function to classify predictors
classify_predictor <- function(predictor) {
  ifelse(grepl("BiologicalMaterial", predictor), "Biological", "Process")
}

# Extract top 5 predictors for each model
get_top5 <- function(model, model_name) {
  importance <- varImp(model, scale = TRUE)$importance
  top5 <- importance %>%
    rownames_to_column("Predictor") %>%
    arrange(desc(Overall)) %>%
    slice_head(n = 5) %>%
    mutate(Model = model_name,
           Type = classify_predictor(Predictor))
  return(top5)
}

# Get top 5 for all nonlinear models
all_top5 <- bind_rows(
  get_top5(knn_model, "KNN"),
  get_top5(nnet_model, "NeuralNet"),
  get_top5(rf_model, "RandomForest"),
  get_top5(mars_model, "MARS"),
  get_top5(svm_model, "SVM")
)

# Display results
print("=== Top 5 Predictors for All Nonlinear Models ===")
print(all_top5)

# Summary: Which variables appear most frequently?
frequent_predictors <- all_top5 %>%
  group_by(Predictor, Type) %>%
  summarise(Appearances = n(), .groups = "drop") %>%
  arrange(desc(Appearances))

print("=== Most Frequently Important Predictors Across Models ===")
print(frequent_predictors)

# Dominance summary by model
dominance <- all_top5 %>%
  group_by(Model, Type) %>%
  summarise(Count = n(), .groups = "drop") %>%
  pivot_wider(names_from = Type, values_from = Count, values_fill = 0) %>%
  mutate(Process_Pct = round(100 * Process / (Process + Biological), 1),
         Biological_Pct = round(100 * Biological / (Process + Biological), 1))

print("=== Biological vs Process Dominance by Model ===")
print(dominance)

# Overall dominance across all models
total_dominance <- all_top5 %>%
  group_by(Type) %>%
  summarise(Total_Appearances = n()) %>%
  mutate(Percentage = round(100 * Total_Appearances / sum(Total_Appearances), 1))

print("=== Overall Dominance (All Nonlinear Models Combined) ===")
print(total_dominance)
```

**ManufacturingProcess32** is universally important — robust finding
across all modeling approaches

**BiologicalMaterial06** is the most important biological variable
(appears in 4 of 5 models)

**MARS** stands out as the only model that completely excludes
biological variables from its top 5

**Model choice matters — NeuralNet and RandomForest** assign more weight
to biological variables than KNN, SVM, or MARS

(c) **Explore the relationships between the top predictors and the
    response for the predictors that are unique to the optimal nonlinear
    regression model. Do these plots reveal intuition about the
    biological or process predictors and their relationship with
    yield?**

```{r}
# ============================================================
# 7.5(c) - Explore Relationships: Top 5 MARS Predictors vs Yield
# ============================================================

library(ggplot2)
library(gridExtra)
library(dplyr)

# Extract top 5 predictors from MARS
mars_importance <- varImp(mars_model, scale = TRUE)
top_predictors <- rownames(mars_importance$importance)[
  order(mars_importance$importance$Overall, decreasing = TRUE)[1:5]
]

# Create plot data
plot_data <- train_data %>%
  select(Yield, all_of(top_predictors))
names(plot_data) <- c("Yield", "Top1", "Top2", "Top3", "Top4", "Top5")

# Plot 1: Top predictor vs Yield
p1 <- ggplot(plot_data, aes(x = Top1, y = Yield)) +
  geom_point(alpha = 0.6, color = "steelblue") +
  geom_smooth(method = "loess", se = TRUE, color = "red") +
  labs(title = paste(top_predictors[1], "vs Yield"),
       subtitle = paste("MARS Importance =", round(mars_importance$importance[top_predictors[1], "Overall"], 1)),
       x = top_predictors[1], y = "Yield (%)") +
  theme_minimal()

# Plot 2: Second predictor vs Yield
p2 <- ggplot(plot_data, aes(x = Top2, y = Yield)) +
  geom_point(alpha = 0.6, color = "darkgreen") +
  geom_smooth(method = "loess", se = TRUE, color = "red") +
  labs(title = paste(top_predictors[2], "vs Yield"),
       subtitle = paste("MARS Importance =", round(mars_importance$importance[top_predictors[2], "Overall"], 1)),
       x = top_predictors[2], y = "Yield (%)") +
  theme_minimal()

# Plot 3: Third predictor vs Yield
p3 <- ggplot(plot_data, aes(x = Top3, y = Yield)) +
  geom_point(alpha = 0.6, color = "purple") +
  geom_smooth(method = "loess", se = TRUE, color = "red") +
  labs(title = paste(top_predictors[3], "vs Yield"),
       subtitle = paste("MARS Importance =", round(mars_importance$importance[top_predictors[3], "Overall"], 1)),
       x = top_predictors[3], y = "Yield (%)") +
  theme_minimal()

# Plot 4: Third predictor vs Yield
p4 <- ggplot(plot_data, aes(x = Top4, y = Yield)) +
  geom_point(alpha = 0.6, color = "purple") +
  geom_smooth(method = "loess", se = TRUE, color = "red") +
  labs(title = paste(top_predictors[4], "vs Yield"),
       subtitle = paste("MARS Importance =", round(mars_importance$importance[top_predictors[4], "Overall"], 1)),
       x = top_predictors[4], y = "Yield (%)") +
  theme_minimal()

# Plot 5: Third predictor vs Yield
p5 <- ggplot(plot_data, aes(x = Top5, y = Yield)) +
  geom_point(alpha = 0.6, color = "purple") +
  geom_smooth(method = "loess", se = TRUE, color = "red") +
  labs(title = paste(top_predictors[5], "vs Yield"),
       subtitle = paste("MARS Importance =", round(mars_importance$importance[top_predictors[5], "Overall"], 1)),
       x = top_predictors[5], y = "Yield (%)") +
  theme_minimal()

# Display all three plots
grid.arrange(p1, p2, p3, ncol = 3)

grid.arrange(p4, p5, ncol = 2)
# Correlation summary
cat("\nCorrelation with Yield:\n")
cat(paste(top_predictors[1], ":", round(cor(plot_data$Top1, plot_data$Yield), 3), "\n"))
cat(paste(top_predictors[2], ":", round(cor(plot_data$Top2, plot_data$Yield), 3), "\n"))
cat(paste(top_predictors[3], ":", round(cor(plot_data$Top3, plot_data$Yield), 3), "\n"))
cat(paste(top_predictors[4], ":", round(cor(plot_data$Top2, plot_data$Yield), 4), "\n"))
cat(paste(top_predictors[5], ":", round(cor(plot_data$Top3, plot_data$Yield), 5), "\n"))

```

**Relationship Findings - Top 5 MARS Predictors vs Yield**

- ManufacturingProcess32 (Importance = 100): Moderate positive
  correlation (r = 0.605) — increasing this process variable improves
  yield
- ManufacturingProcess09 (Importance = 58.0): Moderate positive
  correlation (r = 0.482) — also beneficial but less impactful than
  Process32
- ManufacturingProcess13 (Importance = 29.8): Moderate negative
  correlation (r = -0.478) — higher values are associated with lower
  yield
- ManufacturingProcess01 (Importance = 17.4): Moderate positive
  correlation (r = 0.483) — increasing this process variable improves
  yield
- ManufacturingProcess22 (Importance = 17.4): Moderate negative
  correlation (r = -0.478) — higher values reduce yield, suggesting
  optimal operation at lower levels

**Intuition Revealed**

- All five top predictors are process variables (not biological). To
  maximize yield:

- Increase ManufacturingProcess32, ManufacturingProcess09 and
  ManufacturingProcess01

- Decrease ManufacturingProcess13 and ManufacturingProcess22

*These are actionable levers that can be controlled in real-time during
manufacturing, unlike biological variables which are fixed raw material
properties*

------------------------------------------------------------------------
