---
title: "Homework"
author: "Mehreen Ali Gillani"
date: "2026-05-06"
output:
  pdf_document: default
  html_document: default
---

### Introduction:

Imagine 10000 receipts sitting on your table. Each receipt represents a transaction with items that were purchased. The receipt is a representation of stuff that went into a customer’s basket - and therefore ‘Market Basket Analysis’.

That is exactly what the Groceries Data Set contains: a collection of receipts with each line representing 1 receipt and the items purchased. Each line is called a transaction and each column in a row represents an item.  The data set is attached.

Your assignment is to use R to mine the data for association rules. You should report support, confidence and lift and your top 10 rules by lift. 

Extra credit: do a simple cluster analysis on the data as well.  Use whichever packages you like.  



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)


```

### 1. Load the Groceres dataset

```{r}
# ================================
# Market Basket Analysis & Clustering
# Groceries Dataset
# ================================

# 1. Load required packages and dataset
# install.packages(c("arules", "arulesViz", "factoextra", "cluster", "pheatmap"))
library(arules)        # For association rules
library(arulesViz)     # For visualizing rules (optional)
library(cluster)       # For clustering
library(factoextra)    # For visualizing clusters
library(pheatmap)

data("Groceries")
summary(Groceries)


```
#### Dataset overview:

- This dataset has 9835 transactions (receipts) and 169 items

- Average basket size is 4.4 items, with most baskets containing 1-6 items (median = 3 items). 

- The most frequently purchased item is "whole milk" (appearing in 2,513 baskets, 25.6% of transactions), followed by other vegetables, rolls/buns, soda, and yogurt.

### 2. Mine association rules

```{r}

# Use Apriori algorithm with reasonable thresholds
# (adjust support/confidence for meaningful rules)
rules <- apriori(Groceries, 
                 parameter = list(supp = 0.001,    # support threshold (0.1%)
                                  conf = 0.5,      # confidence threshold (50%)
                                  target = "rules",
                                  minlen = 2))     # at least 2 items per rule

# 4. Sort rules by lift (highest first)
rules_by_lift <- sort(rules, by = "lift", decreasing = TRUE)
```

#### Apriori algorithm Output:

- The Apriori algorithm generated 5,668 association rules using minimum support of 0.001 (at least 9 transactions) and confidence of 0.5. 

- The algorithm successfully processed all 9,835 transactions with 169 items, reducing to 157 items after recoding (excluding items too rare to meet the support threshold). 

- Rules include itemsets from size 2 up to size 6 as specified by maxlen=10.

### 3. Display top 10 rules by lift with support, confidence, lift
```{r}

# ========== TOP 10 RULES BY LIFT ==========
inspect(rules_by_lift[1:10])


```
#### TOP 10 RULES BY LIFT, Observation

- The highest lift rule (18.99) shows that buying "Instant food products" and "soda" together makes customers 19 times more likely to buy "hamburger meat" than buying it randomly. 

- All top 10 rules have lift values above 11, indicating strong positive associations, with 8 of the 10 rules involving dairy products (butter, cream cheese, yogurt) or meat items (hamburger meat, ham). 

- Most rules have low support (0.001-0.002 or 10-19 transactions) but high confidence (50-67%), meaning these are niche but reliable purchasing patterns.


### 4. Extra Credit: Cluster Analysis on Items

```{r}

# Create binary matrix (items x transactions)
item_matrix <- t(as(Groceries, "matrix"))

# Jaccard distance & hierarchical clustering
item_dist <- dist(item_matrix, method = "binary")
hc_items <- hclust(item_dist, method = "average")

# Cut into 5 clusters
clusters <- cutree(hc_items, k = 5)

# Display results
cat("\n========== ITEM CLUSTERS (k=5) ==========\n")
for (i in 1:5) {
  items_in_cluster <- names(clusters[clusters == i])
  cat(paste0("\nCluster ", i, " (", length(items_in_cluster), " items):\n"))
  cat(paste(items_in_cluster[1:min(10, length(items_in_cluster))], collapse = ", "))
  if (length(items_in_cluster) > 10) cat(", ...")
  cat("\n")
}

# bar plot of cluster sizes
cluster_sizes <- table(clusters)

barplot(cluster_sizes, 
        main = "Item Cluster Sizes (k=5)",
        xlab = "Cluster", 
        ylab = "Number of Items",
        col = 2:6,
        names.arg = paste("Cluster", 1:5),
        ylim = c(0, max(cluster_sizes) + 10))

# Add count labels on bars
text(x = 1:5, y = cluster_sizes + 5, 
     labels = cluster_sizes, cex = 1.2)

```


**Interesting observation: cluster 1 has 165 items whereas cluster 2-5 has one item each.** 


**Lets check those items frequency and figure out why its happening?**

```{r}
# Check frequency of singleton items
itemFrequency(Groceries)[c("frozen chicken", "liqueur", 
                            "make up remover", "sound storage medium")]
```


**Why 1 cluster has 165 items, others have 1 each**

**Frequencies of the 4 "singleton" items:**

- frozen chicken → appears in 6 receipts (0.06%)

- liqueur → appears in 9 receipts (0.09%)

- make up remover → appears in 8 receipts (0.08%)
- sound storage medium → appears in 1 receipt (0.01%)
The math: For Jaccard distance:

If an item appears in only 1-9 transactions out of 9835

It almost never co-occurs with any other item

Intersection with any other item = 0 (or very tiny)

Jaccard similarity = 0 → distance = 1 (maximum)

*These items are statistical orphans - they don't cluster because they don't share baskets with others.*

**Filter out rare items:**

```{r}

# Function to get cluster summary with different threshold 
cluster_summary <- function(min_support, n_clusters = 4) {
  # Filter items
  keep_items <- itemFrequency(Groceries) >= min_support
  data_filt <- Groceries[, keep_items]
  
  # Cluster
  mat <- t(as(data_filt, "matrix"))
  dist_mat <- dist(mat, method = "binary")
  hc <- hclust(dist_mat, method = "average")
  clusters <- cutree(hc, k = n_clusters)
  
  # Return clean results
  list(
    threshold = paste0(min_support * 100, "%"),
    transactions = paste0(min_support * 9835, "+"),
    items = ncol(data_filt),
    cluster_sizes = as.vector(table(clusters)),
    clusters = clusters
  )
}

# Test different thresholds
thresholds <- c(0.01, 0.005, 0.001)  # 1%, 0.5%, 0.1%
n_transactions <- c(98, 49, 10)

# output
for(i in 1:3) {
  result <- cluster_summary(thresholds[i])
  cat("Threshold:", result$threshold, 
      "| Min transactions:", n_transactions[i],
      "| Items kept:", result$items, "\n")
  cat("  Cluster sizes:", paste(result$cluster_sizes, collapse = ", "), "\n\n")
}
  

```

#### Key insight:

- 1% threshold (98+ transactions) → Most balanced (81 items in main cluster, small outliers)

- 0.5% or 0.1% → Highly skewed because rare items become singletons

**Recommendation:** Use min_support = 0.005 (49+ transactions) for meaningful clusters without too many outliers.



-----------------------------------------------