---
title: "Homework 5 Exponential smoothing"
author: "Mehreen Ali Gillani"
date: "2026-03-04"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
library(fpp3)

```

8.  **1 Consider the the number of pigs slaughtered in Victoria, available in the aus_livestock dataset.**

##### a. Use the ETS() function to estimate the equivalent model for simple exponential smoothing. Find the optimal values of

Alpha and l theta , and generate forecasts for the next four months.

```{r}
# Load the aus_livestock dataset and filter for Victoria pigs
victoria_pigs <- aus_livestock %>%
  filter(State == "Victoria", Animal == "Pigs") %>%
  select(Month, Count)


# Plot the time series
victoria_pigs %>%
  autoplot(Count) +
  labs(title = "Number of Pigs Slaughtered in Victoria",
       y = "Number of Pigs",
       x = "Month")
```

```{r}
# Fit simple exponential smoothing model
# ETS(A,N,N) represents simple exponential smoothing with additive errors
fit <- victoria_pigs %>%
  model(ETS(Count ~ error("A") + trend("N") + season("N")))

# Display model information including optimal parameters
report(fit)

# Extract the optimal parameters
alpha <- fit %>% 
  tidy() %>%
  filter(term == "alpha") %>%
  pull(estimate)

l0 <- fit %>% 
  tidy() %>%
  filter(term == "initial") %>%
  pull(estimate)

cat("Optimal alpha (smoothing parameter):", alpha, "\n")
cat("Optimal initial level (l0):", l0, "\n")

# Generate forecasts for the next 4 months
forecasts <- fit %>%
  forecast(h = 4)

# Display the forecasts
print("Forecasts for the next 4 months:")
forecasts %>%
  as_tibble() %>%
  select(Month, .mean) %>%
  rename(Forecast = .mean)

# Plot historical data with forecasts
victoria_pigs %>%
  autoplot(Count) +
  autolayer(forecasts, level = c(80, 95)) +  # Specify numeric levels
  labs(title = "Number of Pigs Slaughtered in Victoria with Forecasts",
       y = "Number of Pigs",
       x = "Month")
# Alternative plot if you don't want prediction intervals
victoria_pigs %>%
  autoplot(Count) +
  autolayer(forecasts, level = NULL) +  # Remove intervals completely
  labs(title = "Number of Pigs Slaughtered in Victoria with Forecasts (without intervals)",
       y = "Number of Pigs",
       x = "Month")
```

##### b. Compute a 95% prediction interval for the first forecast using \^y±1.96s where s is the standard deviation of the residuals. Compare your interval with the interval produced by R

```{r}
# Fit SES model and generate forecasts with 95% PI
fit <- victoria_pigs %>%
  model(ETS(Count ~ error("A") + trend("N") + season("N")))

forecasts <- fit %>% forecast(h = 4, level = 95)

# Manual 95% PI using ±1.96s
first_forecast <- forecasts %>% as_tibble() %>% slice(1) %>% pull(.mean)
s <- sd(fit %>% augment() %>% pull(.resid), na.rm = TRUE)

manual_lower <- first_forecast - 1.96 * s
manual_upper <- first_forecast + 1.96 * s

# R's 95% PI using hilo()
r_interval <- forecasts %>% 
  hilo(level = 95) %>% 
  as_tibble() %>% 
  slice(1) %>% 
  pull(`95%`)

# Display comparison
cat("Manual 95% PI: [", manual_lower, ", ", manual_upper, "]\n")
cat("R's 95% PI:     [", r_interval$lower, ", ", r_interval$upper, "]\n")


```

**Both methods produce nearly identical intervals, with R's being slightly wider (by \~16 pigs) due to accounting for parameter uncertainty.**

#### 8.5 Data set global_economy contains the annual Exports from many countries. Select one country to analyse.

a.  Plot the Exports series and discuss the main features of the data.

```{r}
# Select Australia as an example country
aus_exports <- global_economy %>%
  filter(Country == "Australia") %>%
  select(Year, Exports)

# Plot the exports series
aus_exports %>%
  autoplot(Exports) +
  labs(title = "Australian Exports (% of GDP)",
       y = "Exports (% of GDP)",
       x = "Year")

# Summary statistics
aus_exports %>%
  features(Exports, list(mean = mean, sd = sd, min = min, max = max))
```

**Trend:** Generally upward trend from 1960s to present

**Volatility:** Notable fluctuations around major economic events

**Potential cycles:** May show business cycle patterns

**Recent period:** Sharp increase in resources boom (2000s)

**Outliers:** Dips during global financial crisis (2008-09)

**b. Use an ETS(A,N,N) model to forecast the series, and plot the forecasts.**

```{r}
# Filter Australia data
aus_exports <- global_economy %>%
  filter(Country == "Australia") %>%
  select(Year, Exports)

# Fit model
fit_ses <- aus_exports %>%
  model(ETS(Exports ~ error("A") + trend("N") + season("N")))

# Forecast 5 years
fc_ses <- fit_ses %>% forecast(h = 5)

# Plot with forecasts
aus_exports %>%
  autoplot(Exports) +
  autolayer(fc_ses, level = 95) +
  labs(title = "SES Forecast - Australian Exports")

# View parameters
report(fit_ses)

```

**Model:** ETS(A,N,N) with alpha = 0.57 gives moderate weight to recent observations.

**Fit:** Initial level of 13.0% of GDP with residual variance of 1.36, indicating reasonable forecast accuracy.

##### c. Compute the RMSE values for the training data.

```{r}
# Filter Australia data
aus_exports <- global_economy %>%
  filter(Country == "Australia") %>%
  select(Year, Exports)

# Fit model
fit_ses <- aus_exports %>%
  model(ETS(Exports ~ error("A") + trend("N") + season("N")))

# Use accuracy() function
rmse_ses <- accuracy(fit_ses) %>%
  pull(RMSE)

cat("SES RMSE:", round(rmse_ses, 3), "\n")

```

##### d. Compare the results to those from an ETS(A,A,N) model. (Remember that the trended model is using one more parameter than the simpler model.) Discuss the merits of the two forecasting methods for this data set.

```{r}
# Filter Australia data
aus_exports <- global_economy %>%
  filter(Country == "Australia") %>%
  select(Year, Exports)

# Fit Holt's linear trend model
fit_holt <- aus_exports %>%
  model(ETS(Exports ~ error("A") + trend("A") + season("N")))

# View parameters
report(fit_holt)

# Get SINGLE RMSE value for Holt's model
rmse_holt <- accuracy(fit_holt) %>% pull(RMSE)

# Get SINGLE RMSE value for SES model
fit_ses <- aus_exports %>%
  model(ETS(Exports ~ error("A") + trend("N") + season("N")))

rmse_ses <- accuracy(fit_ses) %>% pull(RMSE)

# Display single values
cat("Holt's RMSE:", round(rmse_holt, 3), "\n")
cat("SES RMSE:", round(rmse_ses, 3), "\n")

# Compare AIC values
glance(fit_ses) %>% select(AIC, AICc, BIC)
glance(fit_holt) %>% select(AIC, AICc, BIC)

# Generate forecasts for comparison
fc_ses <- fit_ses %>% forecast(h = 5)
fc_holt <- fit_holt %>% forecast(h = 5)

# Plot comparison
aus_exports %>%
  autoplot(Exports) +
  autolayer(fc_ses, level = 95, color = "blue", alpha = 0.3) +  # Added transparency
  autolayer(fc_holt, level = 95, color = "red", alpha = 0.3) +   # Added transparency
  labs(title = "SES (blue) vs Holt's (red) Forecasts",
       y = "Exports (%)", x = "Year") +
  guides(color = guide_legend(title = "Model")) 
```

**Interpretation**

**RMSE:** Holt's is slightly lower (1.117 vs 1.147), indicating marginally better fit

**AIC/AICc/BIC:** All lower for SES, suggesting the simpler model is preferred when penalizing for extra parameters

**Holt'sbeta:** Very small (0.0001), meaning the trend component adds almost nothing to the model

##### e. Compare the forecasts from both methods. Which do you think is best?

##### Which is better?

**SES is likely better because:**

Similar RMSE with one less parameter (parsimony)

Lower information criteria (AIC, AICc, BIC)

The trend component in Holt's is essentially flat (β ≈ 0)

The data shows an upward trend, but it's not consistent enough to justify the extra parameter in Holt's model.

##### f. Calculate a 95% prediction interval for the first forecast for each model, using the RMSE values and assuming normal errors. Compare your intervals with those produced using R.

```{r}
# Get first forecast values and RMSEs
fc_ses_1st <- fc_ses %>% as_tibble() %>% slice(1) %>% pull(.mean)
fc_holt_1st <- fc_holt %>% as_tibble() %>% slice(1) %>% pull(.mean)

# Manual 95% PIs using RMSE (ŷ ± 1.96 × RMSE)
manual_ses_lower <- fc_ses_1st - 1.96 * rmse_ses
manual_ses_upper <- fc_ses_1st + 1.96 * rmse_ses

manual_holt_lower <- fc_holt_1st - 1.96 * rmse_holt
manual_holt_upper <- fc_holt_1st + 1.96 * rmse_holt

# R's 95% PIs using hilo()
r_ses_pi <- fc_ses %>% hilo(level = 95) %>% as_tibble() %>% slice(1) %>% pull(`95%`)
r_holt_pi <- fc_holt %>% hilo(level = 95) %>% as_tibble() %>% slice(1) %>% pull(`95%`)

# Display comparison
cat("SES Model:\n")
cat("  Manual 95% PI: [", round(manual_ses_lower, 2), ", ", round(manual_ses_upper, 2), "]\n")
cat("  R's 95% PI:     [", round(r_ses_pi$lower, 2), ", ", round(r_ses_pi$upper, 2), "]\n\n")

cat("Holt's Model:\n")
cat("  Manual 95% PI: [", round(manual_holt_lower, 2), ", ", round(manual_holt_upper, 2), "]\n")
cat("  R's 95% PI:     [", round(r_holt_pi$lower, 2), ", ", round(r_holt_pi$upper, 2), "]\n")

```

**Both methods produce nearly identical 95% prediction intervals, with R's intervals being slightly wider (by \~0.04-0.08) due to accounting for parameter uncertainty. Holt's model shows marginally higher forecast values and wider intervals than SES, reflecting its trend component**

##### 8.6 Forecast the Chinese GDP from the `global_economy` data set using an ETS model. Experiment with the various options in the `ETS()` function to see how much the forecasts change with damped trend, or with a Box-Cox transformation. Try to develop an intuition of what each is doing to the forecasts.

```{r}

# Prepare Chinese GDP data
china_gdp <- global_economy %>%
  filter(Country == "China") %>%
  select(Year, GDP)

# 1. Basic ETS (automatic)
fit_auto <- china_gdp %>%
  model(auto = ETS(GDP))
fc_auto <- fit_auto %>% forecast(h = 20)

# 2. ETS with damped trend (CORRECTED SYNTAX)
fit_damped <- china_gdp %>%
  model(damped = ETS(GDP ~ trend("Ad")))  # "Ad" = additive damped trend
fc_damped <- fit_damped %>% forecast(h = 20)

# 3. ETS with Box-Cox transformation
fit_boxcox <- china_gdp %>%
  model(boxcox = ETS(box_cox(GDP, 0) ~ error("A") + trend("A") + season("N")))
fc_boxcox <- fit_boxcox %>% forecast(h = 20)

# Back-transform Box-Cox forecasts
fc_boxcox <- fc_boxcox %>%
  mutate(.mean = exp(.mean))


# Plot with offset to see both lines clearly
china_gdp %>%
  autoplot(GDP) +
  # Plot auto forecast with solid line
  autolayer(fc_auto, .mean, color = "red", linewidth = 1.5, linetype = "solid") +
  # Plot damped forecast with dashed line on top
  autolayer(fc_damped, .mean, color = "blue", linewidth = 1.5, linetype = "dashed") +
  labs(title = "Chinese GDP Forecast Comparison",
       subtitle = "Red=Auto (solid), Blue=Damped (dashed) - Damped shows slightly lower growth",
       y = "GDP", x = "Year") +
  theme_minimal()

# See the difference explicitly
difference <- (fc_auto %>% as_tibble() %>% pull(.mean)) - 
              (fc_damped %>% as_tibble() %>% pull(.mean))

cat("Difference between Auto and Damped forecasts:\n")
cat("Year 1:", round(difference[1]/1e9, 2), "billion USD\n")
cat("Year 20:", round(difference[20]/1e9, 2), "billion USD\n")

# Option 1: Faceted plots (best for comparison)
bind_rows(
  fc_auto %>% as_tibble() %>% mutate(Model = "Auto"),
  fc_damped %>% as_tibble() %>% mutate(Model = "Damped")
) %>%
  ggplot(aes(x = Year, y = .mean, color = Model)) +
  geom_line(linewidth = 1.2) +
  labs(title = "Chinese GDP Forecast Comparison",
       y = "GDP", x = "Year") +
  theme_minimal()

# Option 2: Plot separately to confirm both exist
p1 <- china_gdp %>%
  autoplot(GDP) +
  autolayer(fc_auto, .mean, color = "red", linewidth = 1.5) +
  labs(title = "Auto Forecast Only", y = "GDP", x = "Year")

p2 <- china_gdp %>%
  autoplot(GDP) +
  autolayer(fc_damped, .mean, color = "blue", linewidth = 1.5) +
  labs(title = "Damped Forecast Only", y = "GDP", x = "Year")

library(patchwork)
p1 / p2  # Shows both plots stacked

# View model parameters
report(fit_auto)
report(fit_damped)
report(fit_boxcox)

```

###### Key insights:

-   All models have alpha = 1, giving nearly full weight to most recent observations

-   Damped trend (fi=0.98) slightly reduces long-term growth rate

-   Box-Cox (log transformation) provides dramatically better AIC fit due to scale stabilization

-   Box-Cox model suggests slower trend growth (beta=0.108 vs 0.312) on log scale

**The Box-Cox model has much lower AIC because:**

-   Log transformation stabilized the variance

-   Made the data more "model-friendly"

-   The negative value is fine - AIC is relative, can be negative

##### 8.7 Find an ETS model for the Gas data from aus_production and forecast the next few years. Why is multiplicative seasonality necessary here? Experiment with making the trend damped. Does it improve the forecasts?

```{r}
library(ggtime)

# Load Gas data from aus_production
gas_data <- aus_production %>%
  select(Quarter, Gas)

# Plot the data
gas_data %>%
  autoplot(Gas) +
  labs(title = "Australian Gas Production", y = "Gas production", x = "Quarter")

# 1. ETS with multiplicative seasonality (automatic)
fit_multi <- gas_data %>%
  model(ETS(Gas ~ error("M") + trend("A") + season("M")))
report(fit_multi)

# 2. ETS with additive seasonality (for comparison)
fit_add <- gas_data %>%
  model(ETS(Gas ~ error("A") + trend("A") + season("A")))
report(fit_add)

# 3. ETS with damped trend and multiplicative seasonality
fit_damped <- gas_data %>%
  model(ETS(Gas ~ error("M") + trend("Ad") + season("M")))
report(fit_damped)

# Generate forecasts (next 4 years = 16 quarters)
fc_multi <- fit_multi %>% forecast(h = 16)
fc_add <- fit_add %>% forecast(h = 16)
fc_damped <- fit_damped %>% forecast(h = 16)

# Compare forecasts
gas_data %>%
  autoplot(Gas) +
  autolayer(fc_multi, .mean, color = "red", size = 1) +
  autolayer(fc_add, .mean, color = "blue", size = 1) +
  autolayer(fc_damped, .mean, color = "green", size = 1) +
  labs(title = "Gas Production Forecast Comparison",
       subtitle = "Red=Multiplicative, Blue=Additive, Green=Damped Multiplicative",
       y = "Gas production", x = "Quarter")

# Compare model fit
bind_rows(
  glance(fit_multi) %>% mutate(Model = "Multiplicative"),
  glance(fit_add) %>% mutate(Model = "Additive"),
  glance(fit_damped) %>% mutate(Model = "Damped Multiplicative")
) %>% select(Model, AIC, AICc, BIC)

# Check residuals
fit_multi %>% gg_tsresiduals()
fit_add %>% gg_tsresiduals()
fit_damped %>% gg_tsresiduals()

```

##### Key Findings:

Multiplicative seasonality is essential - Additive model has AIC \~190 points higher, confirming seasonal amplitude grows with trend Damping does NOT improve forecasts - Damped model has higher AIC (+3.1) than undamped multiplicative, suggesting:

-   The trend is consistently linear, not slowing down
-   Adding the damping parameter (φ) overcomplicates without benefit
-   The extra parameter penalty outweighs any fit improvement **Best model: Standard multiplicative seasonality without damped trend**

#### 8.8 Recall your retail time series data (from Exercise 7 in Section 2.10)
**a. Why is multiplicative seasonality necessary for this series?**

```{r}
# Set seed and select random retail series (as per Exercise 7)
set.seed(12345678)
myseries <- aus_retail %>%
  filter(`Series ID` == sample(aus_retail$`Series ID`, 1))

# Why is multiplicative seasonality necessary?
myseries %>%
  autoplot(Turnover) +
  labs(title = "Retail Turnover - Selected Series",
       subtitle = "Seasonal amplitude increases with trend level")
```

**Seasonal amplitude increases with trend level**

##### b. Apply Holt-Winters’ multiplicative method to the data. Experiment with making the trend damped.

```{r}
#  Apply Holt-Winters' multiplicative method
fit_multi <- myseries %>%
  model(multiplicative = ETS(Turnover ~ error("M") + trend("A") + season("M")))
report(fit_multi)

#  Experiment with damped trend
fit_damped <- myseries %>%
  model(damped = ETS(Turnover ~ error("M") + trend("Ad") + season("M")))
report(fit_damped)
```

##### c. Compare the RMSE of the one-step forecasts from the two methods. Which do you prefer?

```{r}
# Compare RMSE of one-step forecasts
accuracy_multi <- accuracy(fit_multi) %>% select(.model, RMSE, MAE, MAPE)
accuracy_damped <- accuracy(fit_damped) %>% select(.model, RMSE, MAE, MAPE)

# Display comparison
bind_rows(accuracy_multi, accuracy_damped) %>%
  arrange(RMSE)

# Calculate which is better
rmse_multi <- accuracy_multi$RMSE
rmse_damped <- accuracy_damped$RMSE

cat("Multiplicative RMSE:", round(rmse_multi, 4), "\n")
cat("Damped RMSE:", round(rmse_damped, 4), "\n")
cat("Difference:", round(rmse_damped - rmse_multi, 4), "\n")

if(rmse_multi < rmse_damped) {
  cat("Prefer MULTIPLICATIVE model (lower RMSE)")
} else {
  cat("Prefer DAMPED model (lower RMSE)")
}

```

Multiplicative RMSE: 0.613 loswer than Damped RMSE: 0.6155

**Prefer MULTIPLICATIVE model (lower RMSE)**

**d. Check that the residuals from the best method look like white noise.**

```{r}
# Check residuals from best method (multiplicative model)
fit_multi <- myseries %>%
  model(multiplicative = ETS(Turnover ~ error("M") + trend("A") + season("M")))

# 1. Visual residual diagnostics
fit_multi %>% gg_tsresiduals()

# 2. Ljung-Box test for autocorrelation (p-value > 0.05 indicates white noise)
ljung_box_result <- fit_multi %>%
  augment() %>%
  features(.resid, ljung_box, lag = 24, dof = 15)  # dof = number of parameters

# Display result
print(ljung_box_result)

# 3. Check if residuals are white noise
if(ljung_box_result$lb_pvalue > 0.05) {
  cat("Residuals look like white noise (p-value =", 
      round(ljung_box_result$lb_pvalue, 3), "> 0.05)")
} else {
  cat("Residuals show significant autocorrelation (p-value =", 
      round(ljung_box_result$lb_pvalue, 3), "< 0.05)")
}

# 4. Additional: Histogram of residuals
fit_multi %>%
  augment() %>%
  ggplot(aes(x = .resid)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "black") +
  labs(title = "Distribution of Residuals", x = "Residuals", y = "Count") +
  theme_minimal()

# 5. ACF plot of residuals
fit_multi %>%
  augment() %>%
  ACF(.resid) %>%
  autoplot() +
  labs(title = "ACF of Residuals")
```

**The Ljung-Box test confirms the residuals are NOT white noise (p-value = 0). This indicates the model hasn't captured all the patterns in the data.**

```{r}
# Try alternative models to improve residuals
fit_improved <- myseries %>%
  model(
    # Try different ETS specifications
    auto = ETS(Turnover),  # Let algorithm choose
    damped = ETS(Turnover ~ error("M") + trend("Ad") + season("M")),
    additive = ETS(Turnover ~ error("A") + trend("A") + season("A"))
  )

# Compare residuals
fit_improved %>%
  augment() %>%
  group_by(.model) %>%
  features(.resid, ljung_box, lag = 24, dof = 8) %>%
  print()

# View which model has highest p-value (closer to white noise)
```

None of the models produce white noise residuals (all p-values \< 0.05). The additive model has the highest p-value (0.00000465), making it the "least bad" option, but still shows significant autocorrelation.

##### e. Now find the test set RMSE, while training the model to the end of 2010. Can you beat the seasonal naïve approach from Exercise 7 in Section 5.11?

```{r}
# Create training set (up to 2010)
myseries_train <- myseries %>%
  filter(year(Month) < 2011)

# Create test set (2011 onwards)
myseries_test <- myseries %>%
  filter(year(Month) >= 2011)

# 1. Seasonal naïve model (benchmark)
fit_snaive <- myseries_train %>%
  model(snaive = SNAIVE(Turnover))

# 2. Best ETS model from previous analysis (additive)
fit_ets <- myseries_train %>%
  model(ets = ETS(Turnover ~ error("A") + trend("A") + season("A")))

# 3. Alternative ETS models
fit_ets_multi <- myseries_train %>%
  model(ets_multi = ETS(Turnover ~ error("M") + trend("A") + season("M")))

fit_ets_damped <- myseries_train %>%
  model(ets_damped = ETS(Turnover ~ error("A") + trend("Ad") + season("A")))

# Generate forecasts for test period
fc_snaive <- fit_snaive %>% forecast(new_data = myseries_test)
fc_ets <- fit_ets %>% forecast(new_data = myseries_test)
fc_ets_multi <- fit_ets_multi %>% forecast(new_data = myseries_test)
fc_ets_damped <- fit_ets_damped %>% forecast(new_data = myseries_test)

# Calculate test set RMSE
accuracy_snaive <- accuracy(fc_snaive, myseries_test) %>% 
  select(.model, RMSE, MAE, MAPE) %>% 
  mutate(Type = "Test Set")

accuracy_ets <- accuracy(fc_ets, myseries_test) %>% 
  select(.model, RMSE, MAE, MAPE) %>% 
  mutate(Type = "Test Set")

accuracy_ets_multi <- accuracy(fc_ets_multi, myseries_test) %>% 
  select(.model, RMSE, MAE, MAPE) %>% 
  mutate(Type = "Test Set")

accuracy_ets_damped <- accuracy(fc_ets_damped, myseries_test) %>% 
  select(.model, RMSE, MAE, MAPE) %>% 
  mutate(Type = "Test Set")

# Combine results
results <- bind_rows(
  accuracy_snaive,
  accuracy_ets,
  accuracy_ets_multi,
  accuracy_ets_damped
) %>%
  arrange(RMSE)

print("Test Set RMSE Comparison:")
print(results)

# Check if ETS beats seasonal naïve
best_ets_rmse <- min(accuracy_ets$RMSE, accuracy_ets_multi$RMSE, accuracy_ets_damped$RMSE)
snaive_rmse <- accuracy_snaive$RMSE

if(best_ets_rmse < snaive_rmse) {
  cat("\nETS beats seasonal naïve! RMSE improvement:", 
      round(snaive_rmse - best_ets_rmse, 3))
} else {
  cat("\nSeasonal naïve performs better. ETS RMSE is higher by:", 
      round(best_ets_rmse - snaive_rmse, 3))
}

# Visual comparison
myseries %>%
  autoplot(Turnover) +
  autolayer(fc_snaive, .mean, color = "red", size = 0.8) +
  autolayer(fc_ets, .mean, color = "blue", size = 0.8) +
  autolayer(fc_ets_multi, .mean, color = "green", size = 0.8) +
  autolayer(fc_ets_damped, .mean, color = "purple", size = 0.8) +
  labs(title = "Forecast Comparison on Test Set (2011+)",
       subtitle = "Red=SNAIVE, Blue=ETS Additive, Green=ETS Multi, Purple=ETS Damped",
       y = "Turnover", x = "Month")

# c) Compare RMSE of one-step forecasts
accuracy_multi <- accuracy(fit_multi) %>% select(.model, RMSE, MAE, MAPE)
accuracy_damped <- accuracy(fit_damped) %>% select(.model, RMSE, MAE, MAPE)

bind_rows(accuracy_multi, accuracy_damped) %>%
  arrange(RMSE)

# Which is preferred? (lower RMSE is better)

# d) Check residuals of best model
best_model <- if_else(
  accuracy_multi$RMSE < accuracy_damped$RMSE,
  "multiplicative",
  "damped"
)


# Generate forecasts to visualize
fc_multi <- fit_multi %>% forecast(h = 24)
fc_damped <- fit_damped %>% forecast(h = 24)

myseries %>%
  autoplot(Turnover) +
  autolayer(fc_multi, .mean, color = "red", size = 1) +
  autolayer(fc_damped, .mean, color = "blue", size = 1) +
  labs(title = "Forecast Comparison",
       subtitle = "Red=Multiplicative, Blue=Damped",
       y = "Turnover", x = "Month")

```

##### Conclusions

**ETS Damped beats seasonal naïve by 0.253 RMSE (16% improvement)**

**Interesting contrast:**

-   Multiplicative model had lower training RMSE (0.613 vs 0.615)

-   But damped model performs better on test data (1.30 vs 1.78)

-   This suggests multiplicative model overfits to training data

-   **Best model for this series:** ETS Damped (lowest test RMSE)

-   **Residuals note:** Still show autocorrelation (p-value ≈ 0), but model generalizes well to unseen data

##### 8.9 For the same retail data, try an STL decomposition applied to the Box-Cox transformed series, followed by ETS on the seasonally adjusted data. How does that compare with your best previous forecasts on the test set?

```{r}

# Training and test sets
myseries_train <- myseries %>% filter(year(Month) < 2011)
myseries_test <- myseries %>% filter(year(Month) >= 2011)

# Approach: STL decomposition with Box-Cox transformation + ETS on seasonally adjusted data

# Find optimal Box-Cox lambda
lambda <- myseries_train %>%
  features(Turnover, features = guerrero) %>%
  pull(lambda_guerrero)

cat("Optimal Box-Cox lambda:", lambda, "\n")

# Apply STL decomposition to Box-Cox transformed series
stl_fit <- myseries_train %>%
  model(
    stl_ets = decomposition_model(
      STL(box_cox(Turnover, lambda) ~ season(window = "periodic")),
      ETS(season_adjust)
    )
  )

# Generate forecasts and back-transform using forecast parameter
fc_stl <- stl_fit %>% 
  forecast(new_data = myseries_test, lambda = lambda)  # Lambda parameter handles back-transform

# Compare with best previous model (ETS Damped)
fit_damped <- myseries_train %>%
  model(ets_damped = ETS(Turnover ~ error("A") + trend("Ad") + season("A")))

fc_damped <- fit_damped %>% forecast(new_data = myseries_test)

# Calculate test set RMSE
accuracy_stl <- accuracy(fc_stl, myseries_test) %>%
  select(.model, RMSE, MAE, MAPE) %>%
  mutate(Model = "STL+ETS")

accuracy_damped <- accuracy(fc_damped, myseries_test) %>%
  select(.model, RMSE, MAE, MAPE) %>%
  mutate(Model = "ETS Damped")

# Add seasonal naïve benchmark
fit_snaive <- myseries_train %>%
  model(snaive = SNAIVE(Turnover))
fc_snaive <- fit_snaive %>% forecast(new_data = myseries_test)
accuracy_snaive <- accuracy(fc_snaive, myseries_test) %>%
  select(.model, RMSE, MAE, MAPE) %>%
  mutate(Model = "Seasonal Naïve")

# Combine results
results <- bind_rows(
  accuracy_stl,
  accuracy_damped,
  accuracy_snaive
) %>%
  select(Model, RMSE, MAE, MAPE) %>%
  arrange(RMSE)

print(results)
```

##### Key Findings

**ETS Damped remains the best model (RMSE = 1.30)**

Beats seasonal naïve by 16%

Beats STL+ETS by 68%

**STL+ETS performs poorly (RMSE = 4.02)**

Almost 3x worse than ETS Damped

Suggests decomposition approach overcomplicates for this series

**Conclusion:** Your original ETS Damped model is the clear winner for this retail data. The STL decomposition with Box-Cox transformation actually harms forecast accuracy, likely due to:

-   Error propagation from multiple steps

-   Difficulty capturing retail dynamics

-   Over-smoothing important patterns

**Key insight:** For retail data with complex seasonal patterns, a single integrated model (ETS Damped) outperforms the decomposed approach because it better captures the interaction between trend, seasonality, and error components. The two-step STL+ETS approach loses these subtle relationships. \_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_
