---
title: "Project1 data_624"
author: "Mehreen Ali Gillani"
Date: "2026-03-25"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    fig_width: 6
    fig_height: 4
    highlight: default
    latex_engine: xelatex        # Better font support
  html_document:
    toc: true                    # Table of contents
    toc_float: true              # Floating TOC
    toc_depth: 3                 # Show headings down to level 3
    number_sections: true        # Number sections
    theme: flatly                # Better theme (flatly, cosmo, darkly)
    highlight: tango             # Code highlighting
    fig_width: 8                 # Figure width in inches
    fig_height: 5                # Figure height in inches
    fig_caption: true            # Figure captions
    code_folding: show           # Show/hide code
    df_print: paged              # Paginated data frames
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# import necessary libraries
library(tidyverse)
library(readxl)
library(lubridate)
library(tseries)
library(knitr)
library(scales)
library(gridExtra)
library(forecast)
library(tsibble)
library(fable)
library(feasts)
```


**This project consists of 3 parts - two required and one bonus and is worth 15% of your grade. The project is due at 11:59 PM on Sunday March 29. I will accept late submissions with a penalty unti the meetup after that when we review some projects.** 

# Part A

**Part A – ATM Forecast, ATM624Data.xlsx In part A, I want you to forecast how much cash is taken out of 4 different ATM machines for May 2010. The data is given in a single file. The variable ‘Cash’ is provided in hundreds of dollars, other than that it is straight forward. I am being somewhat ambiguous on purpose to make this have a little more business feeling. Explain and demonstrate your process, techniques used and not used, and your actual forecast. I am giving you data via an excel file, please provide your written report on your findings, visuals, discussion and your R code via an RPubs link along with the actual.rmd file Also please submit the forecast which you will put in an Excel readable file.**

## Data Loading & Exploration

```{r}
# Load data
file_path <- "ATM624Data.xlsx"
raw <- read_excel(file_path, sheet = "ATM Data")
colnames(raw) <- c("Date", "ATM", "Cash")
raw <- raw %>%
  mutate(Date = as.Date(Date, origin = "1904-01-01"))

glimpse(raw)
```

## Explore Date

```{r}
cat("Date range:", as.character(min(raw$Date, na.rm=TRUE)),
    "to", as.character(max(raw$Date, na.rm=TRUE)), "\n")
cat("Total rows:", nrow(raw), "\n")
cat("ATMs:", paste(unique(na.omit(raw$ATM)), collapse=", "), "\n\n")

# Summary stats
raw %>%
  filter(!is.na(ATM)) %>%
  group_by(ATM) %>%
  summarise(
    N         = n(),
    Mean      = round(mean(Cash, na.rm=TRUE), 1),
    Median    = round(median(Cash, na.rm=TRUE), 1),
    SD        = round(sd(Cash, na.rm=TRUE), 1),
    Min       = round(min(Cash, na.rm=TRUE), 1),
    Max       = round(max(Cash, na.rm=TRUE), 1),
    `NAs`     = sum(is.na(Cash))
  ) %>%
  kable(caption = "Table 1: Summary Statistics by ATM (Cash in $00s)")
```

-   **ATM1** and **ATM2** are conventional machines with consistent
    activity. ATM1 averages \~\$8,390/day; ATM2 \~\$6,260/day.
-   **ATM3** has negligible average withdrawals (0.7 in \$00s = \$70),
    with most days recording zero cash taken out, indicating it may be
    rarely used or potentially out of service.
-   **ATM4** shows an extreme maximum (\$1,091,976) and massive standard
    deviation, suggesting at least one significant outlier.

## Data Issues & Cleaning

### Missing Values

```{r missing-values}
# Identify rows with NA ATM
raw %>%
  filter(!is.na(ATM), is.na(Cash)) %>%
  select(ATM, Date) %>%
  arrange(ATM, Date) %>%
  print()
```

**ATM1:** Missing cash on June 14, 17, 23, 2013 **ATM2:** Missing cash
on June 19, 25, 2013

```{r}
# Quick summary
summary(raw$Date)
range(raw$Date)
```

### Handle Missing value

```{r}
library(zoo)
raw <- raw %>%
  group_by(ATM) %>%           # 1. Process each ATM separately
  arrange(Date) %>%           # 2. Sort by Date within each ATM
  mutate(                      # 3. Create new column
    Cash_filled = ifelse(      # 4. Conditional logic
      is.na(Cash),             # 5. If Cash is missing...
      na.approx(Cash, maxgap = 2, na.rm = FALSE),  # 6. ...interpolate
      Cash                     # 7. ...otherwise keep original value
    )
  ) %>%
  ungroup()  
colSums(is.na(raw))
```

## Training data

```{r}
# Filter data up to May 2, 2014 (last known values)
train_data <- raw %>%
  filter(Date <= as.Date("2014-05-02"))  # Include May 2 as last training point

# Forecast May 3-15, 2014 (13 days)
forecast_Dates <- seq(as.Date("2014-05-03"), as.Date("2014-05-15"), by = "day")
forecast_horizon <- length(forecast_Dates)  # 13 days

# Check
print(forecast_Dates)

print(forecast_horizon)
```

### 1. Outlier ATM4

```{r}
atm4_data <- train_data %>%
  filter(ATM == "ATM4") %>%
  mutate(
    Q1 = quantile(Cash, 0.25, na.rm = TRUE),
    Q3 = quantile(Cash, 0.75, na.rm = TRUE),
    IQR = Q3 - Q1,
    is_outlier = Cash < (Q1 - 1.5 * IQR) | Cash > (Q3 + 1.5 * IQR)
  )

ggplot(atm4_data, aes(x = Date, y = Cash)) +
  geom_line(color = "gray50") +
  geom_point(aes(color = is_outlier), size = 2) +
  scale_color_manual(values = c("black", "red"), 
                     labels = c("Normal", "Outlier")) +
  labs(title = "ATM4 Time Series with Outliers Highlighted",
       x = "Date", y = "Cash ($00s)") +
  theme_minimal()
```

```{r}
# Count and summarize outliers
atm4_data %>%
  summarise(
    Total_Observations = n(),
    Outlier_Count = sum(is_outlier),
    Outlier_Percentage = round(100 * mean(is_outlier), 2),
    Min_Outlier = min(Cash[is_outlier]),
    Max_Outlier = max(Cash[is_outlier])
  )
```

### 2. Handling outliers

```{r}
# Winsorize ATM4 outliers to 99th percentile
train_data_clean <- train_data %>%
  group_by(ATM) %>%
  mutate(
    Cash = ifelse(ATM == "ATM4",
                  {
                    q99 <- quantile(Cash, 0.99, na.rm = TRUE)
                    q01 <- quantile(Cash, 0.01, na.rm = TRUE)
                    pmin(pmax(Cash, q01), q99)
                  },
                  Cash)
  ) %>%
  ungroup()

# Verify outliers are handled
train_data_clean %>%
  filter(ATM == "ATM4") %>%
  summarise(
    Max = max(Cash),
    Min = min(Cash),
    Mean = mean(Cash)
  )
```

**Why Winsorize?** Preserves data structure while reducing extreme
outlier impact on forecasting models. The 3 outliers likely represent
special events (holidays, machine issues) that won't repeat regularly

**Summary:**

-   **Extreme outliers removed:** Max reduced by 86% (from \$1,091,976
    to \$148,810)

-   **More realistic distribution:** Mean dropped from \$47,400 to
    \$44,736

-   **Preserved variation:** Still captures high withdrawal days but
    within reasonable bounds

    ### 3. Handling outliers for all ATMs

```{r}
# Check outliers for all ATMs
train_data_clean %>%
  group_by(ATM) %>%
  summarise(
    Q1 = quantile(Cash, 0.25, na.rm = TRUE),
    Q3 = quantile(Cash, 0.75, na.rm = TRUE),
    IQR = Q3 - Q1,
    Lower_Bound = Q1 - 1.5 * IQR,
    Upper_Bound = Q3 + 1.5 * IQR,
    Min = min(Cash, na.rm = TRUE),
    Max = max(Cash, na.rm = TRUE),
    Outlier_Count = sum(Cash < Lower_Bound | Cash > Upper_Bound, na.rm = TRUE),
    Outlier_Percentage = round(100 * Outlier_Count / n(), 2)
  )

```

### Box_Cox forATM1

```{r}
library(forecast)
train_data_final <- train_data_clean %>%
  group_by(ATM) %>%
  mutate(
    Cash_transformed = ifelse(ATM == "ATM1", 
                              BoxCox(Cash, lambda = BoxCox.lambda(Cash)),
                              Cash)
  ) %>%
  ungroup()
```

**Box-Cox transformation was applied to stabilize variance and make the
time series more normally distributed, which is a key assumption for
many forecasting models like ARIMA. This transformation helps reduce the
impact of the 51 outliers in ATM1 (13.97% of data) and improves forecast
accuracy by making patterns more consistent and predictable.**

```{r}
# Remove row with missing ATM and interpolate NAs
train_data_final <- train_data_final %>%
  filter(!is.na(ATM)) %>%  # Remove the NA row
  group_by(ATM) %>%
  arrange(Date) %>%
  mutate(Cash_transformed = na.approx(Cash_transformed, na.rm = FALSE)) %>%
  ungroup()

# Verify no NAs remain
train_data_final %>%
  filter(Date <= as.Date("2014-05-02")) %>%
  group_by(ATM) %>%
  summarise(
    Total_Rows = n(),
    NA_Count = sum(is.na(Cash_transformed)),
    NA_Percentage = round(100 * NA_Count / n(), 2)
  )
```

## Forecasting 

### Forecasting for ATM1

```{r atm1}
# Prepare data without transformation issues
forecast_results <- list()

# ATM1 - Use simpler approach
atm1_data <- train_data_final %>%
  filter(ATM == "ATM1", Date <= as.Date("2014-05-02")) %>%
  pull(Cash)  # Use original Cash, not transformed

# Remove NAs
atm1_data <- na.omit(atm1_data)

# Create time series
ts_atm1 <- ts(atm1_data, frequency = 7)

# Fit ARIMA on original scale
fit_atm1 <- auto.arima(ts_atm1, seasonal = TRUE)
fc_atm1 <- forecast(fit_atm1, h = 13)

# Compare ARIMA vs ETS for ATM1
fit_ets_atm1 <- ets(ts_atm1)
fit_arima_atm1 <- auto.arima(ts_atm1, seasonal = TRUE)

# Use the better model
if(AIC(fit_arima_atm1) < AIC(fit_ets_atm1)) {
  best_fit <- fit_arima_atm1
  best_name <- "ARIMA"
} else {
  best_fit <- fit_ets_atm1
  best_name <- "ETS"
}

# Print the best model
cat("\n========================================\n")
cat("ATM1 Model Selection Results:\n")
cat("========================================\n")
cat("ETS AIC:", round(AIC(fit_ets_atm1), 2), "\n")
cat("ARIMA AIC:", round(AIC(fit_arima_atm1), 2), "\n")
cat("\n✓ Best model:", best_name, "\n")
cat("AIC difference:", round(abs(AIC(fit_arima_atm1) - AIC(fit_ets_atm1)), 2), "\n")
cat("========================================\n")

fc_best <- forecast(best_fit, h = 13)
```

**Interpretation:**

ARIMA is the superior model for ATM1 with an AIC that is 1,190 points
lower than ETS. This large difference indicates ARIMA captures ATM1's
underlying patterns (weekly seasonality, trends, and autocorrelation)
much more effectively than exponential smoothing.

-   ATM1's 51 outliers (13.97% of data)
-   Weekly withdrawal patterns
-   Day-to-day dependencies in cash usage

### Forecasting for ATM2

```{r atm2}
# ATM2 - Compare ARIMA and ETS only
atm2_data <- train_data_final %>%
  filter(ATM == "ATM2", Date <= as.Date("2014-05-02")) %>%
  pull(Cash) %>% na.omit()
ts_atm2 <- ts(atm2_data, frequency = 7)

# Method 1: ARIMA
fit_arima_atm2 <- auto.arima(ts_atm2, seasonal = TRUE)

# Method 2: ETS (Exponential Smoothing)
fit_ets_atm2 <- ets(ts_atm2)

# Compare models using AIC and accuracy
cat("\n========================================\n")
cat("ATM2 Model Comparison Results:\n")
cat("========================================\n")

# Calculate AIC
cat("\n--- AIC Comparison (lower is better) ---\n")
cat(sprintf("ARIMA AIC: %.2f\n", AIC(fit_arima_atm2)))
cat(sprintf("ETS AIC:   %.2f\n", AIC(fit_ets_atm2)))

# Calculate RMSE
cat("\n--- Training Accuracy (lower is better) ---\n")
rmse_arima <- sqrt(mean(residuals(fit_arima_atm2)^2))
rmse_ets <- sqrt(mean(residuals(fit_ets_atm2)^2))
cat(sprintf("ARIMA RMSE: %.2f\n", rmse_arima))
cat(sprintf("ETS RMSE:   %.2f\n", rmse_ets))

# Select best model based on AIC
if(AIC(fit_arima_atm2) < AIC(fit_ets_atm2)) {
  best_model <- fit_arima_atm2
  best_name <- "ARIMA"
  best_aic <- AIC(fit_arima_atm2)
} else {
  best_model <- fit_ets_atm2
  best_name <- "ETS"
  best_aic <- AIC(fit_ets_atm2)
}

cat("\n========================================\n")
cat("✓ Best model for ATM2:", best_name, "\n")
cat("  Best AIC:", round(best_aic, 2), "\n")
cat("========================================\n")

# Generate forecasts
fc_arima <- forecast(fit_arima_atm2, h = 13)
fc_ets <- forecast(fit_ets_atm2, h = 13)

# Display forecast comparison
cat("\n--- Forecast Comparison (May 3-15, 2014) ---\n")
cat("Date       | ARIMA | ETS | Selected\n")
cat("----------------------------------------\n")

forecast_dates <- seq(as.Date("2014-05-03"), as.Date("2014-05-15"), by = "day")

for(i in 1:13) {
  cat(format(forecast_dates[i], "%Y-%m-%d"), "|")
  cat(sprintf(" %5.1f |", fc_arima$mean[i]))
  cat(sprintf(" %5.1f |", fc_ets$mean[i]))
  
  if(best_name == "ARIMA") {
    cat(sprintf(" %5.1f", fc_arima$mean[i]))
  } else {
    cat(sprintf(" %5.1f", fc_ets$mean[i]))
  }
  cat("\n")
}

# Use the best model's forecast
if(best_name == "ARIMA") {
  fc_atm2 <- fc_arima
} else {
  fc_atm2 <- fc_ets
}

cat("\n========================================\n")
cat("Final forecast using:", best_name, "\n")
cat("========================================\n")

# Summary of selected forecast
summary_fc <- data.frame(
  Date = forecast_dates,
  Forecast_00s = round(fc_atm2$mean, 2),
  Lower_80 = round(fc_atm2$lower[,1], 2),
  Upper_80 = round(fc_atm2$upper[,1], 2)
)
print(summary_fc)

cat("\nTotal forecast for May 3-15, 2014:", 
    round(sum(fc_atm2$mean), 0), "($00s)\n")
cat("Total in dollars: $", 
    format(round(sum(fc_atm2$mean) * 100, 0), big.mark = ","), "\n")
```

**Why ARIMA Wins:**

-   **Best fit:** Lowest AIC indicates ARIMA best captures ATM2's
    underlying patterns

-   **Most accurate:** Lowest RMSE on training data

-   **Consistent forecasts:** Similar to other methods but with better
    statistical properties

-   **Handles weekly seasonality well:** Frequency = 7 days

### Forecasting for ATM3

```{r atm3}
# ATM3 - Why simple mean is best for zero-inflated data
atm3_data <- train_data_final %>%
  filter(ATM == "ATM3", Date <= as.Date("2014-05-02")) %>%
  pull(Cash)

# Remove any NAs
atm3_data <- na.omit(atm3_data)

# Check distribution
cat("ATM3 Data Distribution:\n")
cat("Total days:", length(atm3_data), "\n")
cat("Days with zero:", sum(atm3_data == 0), 
    paste0("(", round(100 * sum(atm3_data == 0)/length(atm3_data), 1), "%)\n"))
cat("Days with >0:", sum(atm3_data > 0), "\n")
cat("Maximum value:", max(atm3_data), "\n\n")

# Use historical mean (accounts for zeros and occasional spikes)
atm3_mean <- mean(atm3_data, na.rm = TRUE)
atm3_sd <- sd(atm3_data, na.rm = TRUE)

cat("Historical mean:", round(atm3_mean, 2), "($00s)\n")
cat("Historical SD:", round(atm3_sd, 2), "\n\n")

# Simple mean forecast is most appropriate because:
# 1. No seasonal pattern exists (only 1 non-zero day)
# 2. ARIMA/ETS would overfit to noise
# 3. Mean represents expected daily average including zeros

atm3_forecast <- rep(atm3_mean, 13)

# Confidence intervals using historical SD
atm3_lower_80 <- pmax(atm3_mean - 1.28 * atm3_sd, 0)
atm3_upper_80 <- atm3_mean + 1.28 * atm3_sd

cat("13-Day Forecast:\n")
cat("Total:", round(sum(atm3_forecast), 2), "($00s)\n")
cat("Total in dollars: $", format(round(sum(atm3_forecast) * 100, 0), big.mark = ","), "\n")
```

**Result:**

ATM3 forecast: \$910 total for May 3-15, 2014 This correctly reflects
that ATM3 is rarely used

**Why ARIMA and ETS are NOT Used for ATM3:**

-   **Zero-Inflated Data**

ATM3 has 99% zeros with occasional small spikes. The historical mean is
only 0.7 (\$70 per day). ARIMA and ETS models are designed for
continuous, non-zero data with patterns - they fail dramatically with
this type of data.

### Forecasting for ATM4

```{r atm4}
# ATM4 - Compare ARIMA and ETS
atm4_data <- train_data_final %>%
  filter(ATM == "ATM4", Date <= as.Date("2014-05-02")) %>%
  pull(Cash) %>% na.omit()
ts_atm4 <- ts(atm4_data, frequency = 7)

# Method 1: ARIMA
fit_arima_atm4 <- auto.arima(ts_atm4, seasonal = TRUE)

# Method 2: ETS (Exponential Smoothing)
fit_ets_atm4 <- ets(ts_atm4)

# Compare models
cat("\n========================================\n")
cat("ATM4 Model Comparison Results:\n")
cat("========================================\n")

# Calculate AIC
cat("\n--- AIC Comparison (lower is better) ---\n")
cat(sprintf("ARIMA AIC: %.2f\n", AIC(fit_arima_atm4)))
cat(sprintf("ETS AIC:   %.2f\n", AIC(fit_ets_atm4)))

# Calculate RMSE
cat("\n--- Training Accuracy (lower is better) ---\n")
rmse_arima <- sqrt(mean(residuals(fit_arima_atm4)^2))
rmse_ets <- sqrt(mean(residuals(fit_ets_atm4)^2))
cat(sprintf("ARIMA RMSE: %.2f\n", rmse_arima))
cat(sprintf("ETS RMSE:   %.2f\n", rmse_ets))

# Select best model based on AIC
if(AIC(fit_arima_atm4) < AIC(fit_ets_atm4)) {
  best_model <- fit_arima_atm4
  best_name <- "ARIMA"
  best_aic <- AIC(fit_arima_atm4)
} else {
  best_model <- fit_ets_atm4
  best_name <- "ETS"
  best_aic <- AIC(fit_ets_atm4)
}

cat("\n========================================\n")
cat("✓ Best model for ATM4:", best_name, "\n")
cat("  Best AIC:", round(best_aic, 2), "\n")
cat("========================================\n")

# Generate forecasts
fc_arima <- forecast(fit_arima_atm4, h = 13)
fc_ets <- forecast(fit_ets_atm4, h = 13)

# Display forecast comparison for first 5 days
cat("\n--- Forecast Comparison (First 5 days) ---\n")
cat("Date       | ARIMA | ETS | Selected\n")
cat("----------------------------------------\n")

forecast_dates <- seq(as.Date("2014-05-03"), as.Date("2014-05-15"), by = "day")

for(i in 1:5) {
  cat(format(forecast_dates[i], "%Y-%m-%d"), "|")
  cat(sprintf(" %6.1f |", fc_arima$mean[i]))
  cat(sprintf(" %6.1f |", fc_ets$mean[i]))
  
  if(best_name == "ARIMA") {
    cat(sprintf(" %6.1f", fc_arima$mean[i]))
  } else {
    cat(sprintf(" %6.1f", fc_ets$mean[i]))
  }
  cat("\n")
}

# Use the best model's forecast
if(best_name == "ARIMA") {
  fc_atm4 <- fc_arima
} else {
  fc_atm4 <- fc_ets
}

cat("\n========================================\n")
cat("Final forecast using:", best_name, "\n")
cat("========================================\n")

# Summary of selected forecast
summary_fc <- data.frame(
  Date = forecast_dates,
  Forecast_00s = round(fc_atm4$mean, 2),
  Lower_80 = round(fc_atm4$lower[,1], 2),
  Upper_80 = round(fc_atm4$upper[,1], 2)
)
print(summary_fc)

cat("\nTotal forecast for May 3-15, 2014:", 
    round(sum(fc_atm4$mean), 0), "($00s)\n")
cat("Total in dollars: $", 
    format(round(sum(fc_atm4$mean) * 100, 0), big.mark = ","), "\n")
```

### ATM4 Forecast Results Summary

-   ARIMA outperforms ETS with AIC of 5,311 vs 6,407 (1,096 point
    difference), indicating significantly better model fit

-   **Mixed accuracy metrics:** ARIMA has slightly higher RMSE (341.8 vs
    330.1) but is preferred based on AIC which penalizes model
    complexity **F**

-   **orecast total:** \$540,190 for May 3-15, 2014 (13 days), averaging
    \$41,553 per day

-   **Daily pattern:** Forecast shows variability between \$330-\$505
    (\$00s), reflecting ATM4's historical volatility **Wide confidence
    intervals:** Lower bounds negative (capped at \$0 in actual
    forecasts) and upper bounds up to \$949 (\$00s), indicating high
    uncertainty due to extreme historical outliers

-   **Model selection:** ARIMA chosen over ETS despite slightly higher
    RMSE because it better balances fit with parsimony (fewer
    parameters)

## Combining and visualizing Results:

```{r combine results}
# Create forecast dataframe
forecast_Dates <- seq(as.Date("2014-05-03"), as.Date("2014-05-15"), by = "day")

all_forecasts_fixed <- bind_rows(
  data.frame(Date = forecast_Dates, ATM = "ATM1", 
             Forecast = as.numeric(fc_atm1$mean),
             Lower_80 = as.numeric(fc_atm1$lower[,1]),
             Upper_80 = as.numeric(fc_atm1$upper[,1])),
  data.frame(Date = forecast_Dates, ATM = "ATM2",
             Forecast = as.numeric(fc_atm2$mean),
             Lower_80 = as.numeric(fc_atm2$lower[,1]),
             Upper_80 = as.numeric(fc_atm2$upper[,1])),
  # Create forecast dataframe for ATM3
  data.frame(Date = forecast_Dates, ATM = "ATM3",
             Forecast = atm3_forecast,
             Lower_80 = rep(atm3_lower_80, 13),
             Upper_80 = rep(atm3_upper_80, 13)),
  data.frame(Date = forecast_Dates, ATM = "ATM4",
             Forecast = as.numeric(fc_atm4$mean),
             Lower_80 = as.numeric(fc_atm4$lower[,1]),
             Upper_80 = as.numeric(fc_atm4$upper[,1]))
)

# Ensure non-negative
all_forecasts_fixed <- all_forecasts_fixed %>%
  mutate(across(c(Forecast, Lower_80, Upper_80), ~ pmax(., 0)))
```


```{r  atm-forecast-plots, fig.width=10, fig.height=6, out.width="100%"}
# Plot forecasts
for(atm in c("ATM1", "ATM2", "ATM3", "ATM4")) {
  
  # Historical data
  hist_data <- train_data_final %>%
    filter(ATM == atm, Date <= as.Date("2014-05-02")) %>%
    select(Date, Cash)
  
  # Forecast data
  fc_data <- all_forecasts_fixed %>% filter(ATM == atm)
  
  p <- ggplot() +
    geom_line(data = tail(hist_data, 90), 
              aes(x = Date, y = Cash), 
              color = "black", linewidth = 0.8) +
    geom_line(data = fc_data, 
              aes(x = Date, y = Forecast), 
              color = "blue", linewidth = 1.2) +
    geom_ribbon(data = fc_data, 
                aes(x = Date, ymin = Lower_80, ymax = Upper_80), 
                fill = "blue", alpha = 0.2) +
    labs(title = paste(atm, "- May 3-15, 2014 Cash Withdrawal Forecast"),
         x = "Date", y = "Cash ($00s)") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),  # Rotate x-axis labels
      plot.margin = margin(1, 1, 1, 1, "cm")
      ) # Add margins)
  
  print(p)
}
```


```{r  summary}
# Summary
summary_fixed <- all_forecasts_fixed %>%
  group_by(ATM) %>%
  summarise(
    Total_Forecast = sum(Forecast),
    Average_Daily = mean(Forecast),
    Max_Daily = max(Forecast),
    Min_Daily = min(Forecast),
    CI_Width_Avg = mean(Upper_80 - Lower_80)
  ) %>%
  mutate(Total_Dollars = Total_Forecast * 100)

print(summary_fixed)

# Total
total_fixed <- sum(summary_fixed$Total_Dollars)
cat("\nTotal forecasted cash: $", format(total_fixed, big.mark = ","), "\n")
```

## Discussion & Business Recommendations

### Model Performance Summary

-   ARIMA consistently outperformed ETS across ATM1, ATM2, and ATM4,
    with AIC improvements of 1,190–1,600 points, confirming its
    suitability for cash withdrawal forecasting with weekly seasonality
-   ATM3 required a simple historical mean approach due to zero-inflated
    data (99% zeros), where complex time series models would produce
    unrealistic forecasts
-   Forecast accuracy varied by ATM: ATM2 showed highest reliability
    (RMSE 26.1), while ATM4 exhibited high volatility (RMSE 341.8) due
    to extreme outliers up to \$1.1M

### **Key Findings**

-   Total forecasted cash for May 3-15, 2014: \$715,384
-   ATM4 dominates withdrawals (76% of total), requiring focused cash
    management
-   ATM3 is underutilized (\$937 total), suggesting potential for
    decommissioning

### **Business Recommendations**

1.  **Optimize cash replenishment schedules** using ARIMA forecasts for
    ATM1, ATM2, and ATM4 to reduce idle cash and prevent shortages
2.  **Review ATM4 usage patterns** - investigate causes of extreme
    volatility (business district? special events?) and consider dynamic
    replenishment
3.  **Evaluate ATM3 viability** - assess if operational costs justify
    continued service given minimal usage
4.  **Implement regular forecast validation** - compare actual May 2014
    withdrawals against forecasts to refine models
5.  **Monitor confidence intervals**- use 80% prediction intervals for
    operational planning, 95% for risk management

# Part B 

**Part B – Forecasting Power, ResidentialCustomerForecastLoad-624.xlsx.Part B consists of a simple dataset of residential power usage forJanuary 1998 until December 2013. Your assignment is to model these data and a monthly forecast for 2014. The data is given in a single file. The variable ‘KWH’ is power consumption in Kilowatt hours, the rest is straight forward. Add this to your existing files above.**

## Load Dataset

```{r}
# Read the data
file_path <- "ResidentialCustomerForecastLoad-624.xlsx"
power_data <- read_excel(file_path)

# Set column names
colnames(power_data) <- c("Sequence", "Date", "KWH")

# View structure
glimpse(power_data)
head(power_data)
```

### Extract Proper Date

```{r}
# Extract year and month from the "Date" column (e.g., "1998-Jan")
power_data <- power_data %>%
  mutate(
    # Extract year
    Year = as.numeric(substr(Date, 1, 4)),
    # Extract month abbreviation
    Month_Abbr = substr(Date, 6, 8),
    # Convert month abbreviation to number
    Month = match(Month_Abbr, month.abb),
    # Create proper date (first day of each month)
    Proper_Date = as.Date(paste(Year, Month, "01", sep = "-"))
  ) %>%
  filter(!is.na(KWH))  # Remove any NA values

# Check results
head(power_data[, c("Proper_Date", "Year", "Month", "KWH")])
```

## Date Range

```{r}
# Check date range
cat("Date range:", min(power_data$Proper_Date), "to", max(power_data$Proper_Date), "\n")
cat("Total observations:", nrow(power_data), "\n")

# Check if all months are present
power_data %>%
  group_by(Year) %>%
  summarise(Months = n_distinct(Month)) %>%
  print()
```

## Create Time Series

```{r}
# Create monthly time series
ts_power <- ts(power_data$KWH, 
               start = c(1998, 1), 
               frequency = 12)

# Plot time series
plot(ts_power, 
     main = "Residential Power Usage (Jan 1998 - Dec 2013)",
     xlab = "Year", 
     ylab = "KWH (Kilowatt hours)",
     col = "steelblue",
     lwd = 2)
```

## Visualize Pattern

```{r}
# Method 4: Boxplots to show monthly distribution
power_data %>%
  mutate(Month = month(Proper_Date, label = TRUE)) %>%
  ggplot(aes(x = Month, y = KWH)) +
  geom_boxplot(fill = "steelblue", alpha = 0.6) +
  labs(title = "Monthly Distribution of Power Usage",
       x = "Month", 
       y = "KWH") +
  theme_minimal()
```

## Fit Models and Forecast

```{r}
# Fit ARIMA
fit_arima <- auto.arima(ts_power, seasonal = TRUE)
fc_arima <- forecast(fit_arima, h = 12)

# Fit ETS
fit_ets <- ets(ts_power)
fc_ets <- forecast(fit_ets, h = 12)

# Compare models
cat("\n========================================\n")
cat("Model Comparison:\n")
cat("========================================\n")
cat("ARIMA AIC:", round(AIC(fit_arima), 2), "\n")
cat("ETS AIC:  ", round(AIC(fit_ets), 2), "\n")

# Select best
if(AIC(fit_arima) < AIC(fit_ets)) {
  best_model <- fit_arima
  best_name <- "ARIMA"
} else {
  best_model <- fit_ets
  best_name <- "ETS"
}
cat("\n✓ Best model:", best_name, "\n")

# Generate forecast
forecast_2014 <- forecast(best_model, h = 12)

```

## Create 2014 Forecast Table

```{r}
# Create forecast dates
forecast_months <- seq(as.Date("2014-01-01"), as.Date("2014-12-01"), by = "month")

forecast_results <- data.frame(
  Month = format(forecast_months, "%B %Y"),
  Month_Abbr = month.abb,
  Forecast_KWH = round(forecast_2014$mean, 0),
  Lower_80 = round(forecast_2014$lower[,1], 0),
  Upper_80 = round(forecast_2014$upper[,1], 0),
  Lower_95 = round(forecast_2014$lower[,2], 0),
  Upper_95 = round(forecast_2014$upper[,2], 0)
)

print(forecast_results)
```

## Visualize Forecast

```{r}
# Plot forecast
autoplot(forecast_2014) +
  labs(title = paste("Power Usage Forecast for 2014 (", best_name, " Model)", sep=""),
       x = "Year", 
       y = "KWH (Kilowatt hours)") +
  theme_minimal()

# Combined plot with recent history
power_data_recent <- power_data %>%
  filter(Proper_Date >= as.Date("2010-01-01"))

forecast_plot_data <- data.frame(
  Date = forecast_months,
  KWH = forecast_2014$mean,
  Type = "Forecast"
)

ggplot() +
  geom_line(data = power_data_recent, 
            aes(x = Proper_Date, y = KWH), 
            color = "black", size = 0.8) +
  geom_line(data = forecast_plot_data, 
            aes(x = Date, y = KWH), 
            color = "blue", size = 1.2) +
  geom_ribbon(data = forecast_results, 
              aes(x = forecast_months, 
                  ymin = Lower_80, ymax = Upper_80), 
              fill = "blue", alpha = 0.2) +
  labs(title = "Power Usage Forecast for 2014",
       x = "Date", 
       y = "KWH") +
  scale_x_date(date_breaks = "6 months", date_labels = "%b %Y") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Summary Statistics:

```{r}
# Summary
summary_2014 <- forecast_results %>%
  summarise(
    Total_Annual = sum(Forecast_KWH),
    Average_Monthly = mean(Forecast_KWH),
    Peak_Month = Month[which.max(Forecast_KWH)],
    Peak_Value = max(Forecast_KWH),
    Lowest_Month = Month[which.min(Forecast_KWH)],
    Lowest_Value = min(Forecast_KWH)
  )

cat("\n========================================\n")
cat("2014 FORECAST SUMMARY\n")
cat("========================================\n")
cat("Total annual usage:", format(summary_2014$Total_Annual, big.mark = ","), "KWH\n")
cat("Average monthly:", format(round(summary_2014$Average_Monthly, 0), big.mark = ","), "KWH\n")
cat("Peak month:", summary_2014$Peak_Month, "with", 
    format(summary_2014$Peak_Value, big.mark = ","), "KWH\n")
cat("Lowest month:", summary_2014$Lowest_Month, "with",
    format(summary_2014$Lowest_Value, big.mark = ","), "KWH\n")

```

## Residential Power Usage Forecast Results

### **Model Performance**

-   **ARIMA selected as best model (AIC: 5,455 vs ETS: 6,303)**,
    confirming superior fit for monthly seasonal patterns 2014 Forecast
    Summary

-   **Total annual usage:** 91,313,308 KWH

-   **Average monthly:** 7,609,442 KWH

-   **Peak month:** January 2014 with 10,104,745 KWH (winter heating
    demand)

-   **Lowest month:** October 2014 with 5,860,807 KWH (fall moderate
    weather)

**Seasonal Pattern**

-   **Winter peak:** January (10.1M KWH)

-   **Summer peak:** August (9.7M KWH)

-   **Spring/Fall troughs:** May (5.9M) and October (5.9M)

### Business Recommendations

Ensure capacity for January peak (10.1M KWH) - highest demand month
Schedule maintenance in spring (April-May) during lowest usage periods
Validate forecasts against actual 2014 consumption to refine models

# Part C

**Part C – BONUS, optional (part or all), Waterflow_Pipe1.xlsx and Waterflow_Pipe2.xlsx. Part C consists of two data sets. These are simple 2 columns sets, however they have different time stamps. Your optional assignment is to time-base sequence the data and aggregate based on hour (example of what this looks like, follows). Note for multiple recordings within an hour, take the mean. Then to determine if the data is stationary and can it be forecast. If so, provide a week forward forecast and present results via Rpubs and .rmd and the forecast in an Excel readable file.**

## Load Datafiles:

```{r}
# Read both datasets
pipe1 <- read_excel("Waterflow_Pipe1.xlsx")
pipe2 <- read_excel("Waterflow_Pipe2.xlsx")

# Check structure
cat("===== PIPE 1 DATA =====\n")
glimpse(pipe1)
print(head(pipe1))

cat("\n===== PIPE 2 DATA =====\n")
glimpse(pipe2)
print(head(pipe2))

```

## EDA

```{r}
# Rename columns if needed (assuming first column is timestamp, second is flow)
colnames(pipe1) <- c("Timestamp", "Flow")
colnames(pipe2) <- c("Timestamp", "Flow")

# Convert Excel numeric dates to proper datetime
# Excel origin for Windows: 1899-12-30
pipe1$Timestamp <- as.POSIXct(pipe1$Timestamp * 86400, origin = "1899-12-30")
pipe2$Timestamp <- as.POSIXct(pipe2$Timestamp * 86400, origin = "1899-12-30")

# Check conversion
cat("\n===== PIPE 1 CONVERTED TIMESTAMPS =====\n")
print(head(pipe1))
cat("\nDate range:", min(pipe1$Timestamp), "to", max(pipe1$Timestamp), "\n")

cat("\n===== PIPE 2 CONVERTED TIMESTAMPS =====\n")
print(head(pipe2))
cat("\nDate range:", min(pipe2$Timestamp), "to", max(pipe2$Timestamp), "\n")
```

## Aggregate by Hour (Take Mean)

```{r}
# Aggregate Pipe1 by hour
pipe1_hourly <- pipe1 %>%
  mutate(
    Hour = floor_date(Timestamp, unit = "hour")  # Round down to nearest hour
  ) %>%
  group_by(Hour) %>%
  summarise(
    Flow = mean(Flow, na.rm = TRUE),
    Count = n(),  # Optional: track how many readings per hour
    .groups = 'drop'
  ) %>%
  arrange(Hour)

# Aggregate Pipe2 by hour
pipe2_hourly <- pipe2 %>%
  mutate(
    Hour = floor_date(Timestamp, unit = "hour")
  ) %>%
  group_by(Hour) %>%
  summarise(
    Flow = mean(Flow, na.rm = TRUE),
    Count = n(),
    .groups = 'drop'
  ) %>%
  arrange(Hour)

# Check results
cat("===== PIPE 1 HOURLY AGGREGATION =====\n")
print(head(pipe1_hourly))
cat("\nTotal hourly records:", nrow(pipe1_hourly), "\n")
cat("Hour range:", min(pipe1_hourly$Hour), "to", max(pipe1_hourly$Hour), "\n")

cat("\n===== PIPE 2 HOURLY AGGREGATION =====\n")
print(head(pipe2_hourly))
cat("\nTotal hourly records:", nrow(pipe2_hourly), "\n")
cat("Hour range:", min(pipe2_hourly$Hour), "to", max(pipe2_hourly$Hour), "\n")

```

## Combine Both Pipes

```{r}
# Combine Pipe1 and Pipe2 by hour
combined_flow <- pipe1_hourly %>%
  rename(Flow_Pipe1 = Flow, Count_Pipe1 = Count) %>%
  full_join(pipe2_hourly %>% rename(Flow_Pipe2 = Flow, Count_Pipe2 = Count), by = "Hour") %>%
  mutate(
    # Fill NA with 0 for hours with no data from either pipe
    Flow_Pipe1 = ifelse(is.na(Flow_Pipe1), 0, Flow_Pipe1),
    Flow_Pipe2 = ifelse(is.na(Flow_Pipe2), 0, Flow_Pipe2),
    Count_Pipe1 = ifelse(is.na(Count_Pipe1), 0, Count_Pipe1),
    Count_Pipe2 = ifelse(is.na(Count_Pipe2), 0, Count_Pipe2),
    Total_Flow = Flow_Pipe1 + Flow_Pipe2
  ) %>%
  arrange(Hour)

# View combined results
cat("===== COMBINED HOURLY FLOW =====\n")
print(head(combined_flow, 10))
cat("\nTotal hourly records:", nrow(combined_flow), "\n")
cat("Date range:", min(combined_flow$Hour), "to", max(combined_flow$Hour), "\n")
cat("Total flow range:", min(combined_flow$Total_Flow), "to", max(combined_flow$Total_Flow), "\n")
```

## Create Time Series and Visualize

```{r}
# Create time series object (hourly data with daily seasonality)
ts_combined <- ts(combined_flow$Total_Flow, 
                  frequency = 24)  # 24 hours in a day


# plot with ggplot for better visualization
ggplot(combined_flow, aes(x = Hour, y = Total_Flow)) +
  geom_line(color = "darkgreen", size = 0.6) +
  labs(title = "Combined Water Flow - Hourly Data",
       x = "Date/Time", 
       y = "Total Flow Rate") +
  scale_x_datetime(date_breaks = "2 days", date_labels = "%b %d") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

## Test for stationarity

```{r}
# Create time series object
ts_combined <- ts(combined_flow$Total_Flow, frequency = 24)

# 1. Augmented Dickey-Fuller Test (ADF)
# Null hypothesis: data is non-stationary
adf_test <- adf.test(ts_combined, alternative = "stationary")

cat("========================================\\n")
cat("STATIONARITY TESTS\\n")
cat("========================================\\n")

cat("\\n1. Augmented Dickey-Fuller Test:\\n")
cat(" p-value:", adf_test$p.value, "\\n")
cat(" Conclusion:", ifelse(adf_test$p.value < 0.05,
"Reject null - Data is STATIONARY",
"Fail to reject null - Data is NON-STATIONARY"), "\\n")

# 2. KPSS Test (Kwiatkowski-Phillips-Schmidt-Shin)
# Null hypothesis: data is stationary
kpss_test <- kpss.test(ts_combined)

cat("\\n2. KPSS Test:\\n")
cat(" p-value:", kpss_test$p.value, "\\n")
cat(" Conclusion:", ifelse(kpss_test$p.value > 0.05,
"Fail to reject null - Data is STATIONARY",
"Reject null - Data is NON-STATIONARY"), "\\n")

# 3. Visual inspection - ACF plot
par(mfrow = c(1,1))
acf(ts_combined, main = "Autocorrelation Function (ACF)")
par(mfrow = c(1,1))

# Final conclusion
is_stationary <- (adf_test$p.value < 0.05) & (kpss_test$p.value > 0.05)
cat("\\n========================================\\n")
cat("FINAL CONCLUSION:\\n")
cat("========================================\\n")
cat("Data is", ifelse(is_stationary, "STATIONARY ✓", "NON-STATIONARY
✗"), "\\n")
cat("Can be forecast:", ifelse(is_stationary, "YES", "NO - needs
differencing"), "\\n")
```

-   ADF says STATIONARY (p \< 0.05)
-   KPSS says NON-STATIONARY (p \< 0.05)
-   **Visual inspection:** ACF plot shows slow decay = non-stationary

When tests conflict, we trust KPSS or conclude data needs differencing.
Let's proceed.

## Seasonal differencing

```{r}
# Apply seasonal differencing (lag = 24 for daily pattern)
ts_diff <- diff(ts_combined, lag = 24)

# Test if differencing made it stationary
adf_diff <- adf.test(ts_diff, alternative = "stationary")
kpss_diff <- kpss.test(ts_diff)

cat("========================================\n")
cat("AFTER SEASONAL DIFFERENCING (lag=24)\n")
cat("========================================\n")
cat("ADF p-value:", adf_diff$p.value, "\n")
cat("KPSS p-value:", kpss_diff$p.value, "\n")

is_stationary <- (adf_diff$p.value < 0.05) & (kpss_diff$p.value > 0.05)
cat("\nData is now", ifelse(is_stationary, "STATIONARY ✓", "STILL NON-STATIONARY"), "\n")

# Visualize differenced data
par(mfrow = c(2,1))
plot(ts_diff, main = "Differenced Flow (lag=24)", col = "blue")
acf(ts_diff, main = "ACF of Differenced Data")
par(mfrow = c(1,1))

# Prepare data for forecasting
forecast_data <- ts_diff
cat("\n✓ Data ready for forecasting\n")

```

## Fit ARIMA Model and Forecast One Week

```{r}
 # Fit ARIMA model on stationary data
fit_arima <- auto.arima(ts_diff, seasonal = FALSE)
cat("========================================\n")
cat("ARIMA MODEL SELECTION\n")
cat("========================================\n")
print(summary(fit_arima))

# Generate forecast for 168 hours (7 days)
forecast_hours <- 168
fc <- forecast(fit_arima, h = forecast_hours)

# Reverse the seasonal differencing to get actual forecasts
# Get the last 24 original values to reconstruct
last_original <- tail(ts_combined, 24)

# Build back the forecasted values
fc_mean <- numeric(forecast_hours)
fc_lower_80 <- numeric(forecast_hours)
fc_upper_80 <- numeric(forecast_hours)

for(i in 1:forecast_hours) {
  if(i <= 24) {
    fc_mean[i] <- last_original[i] + fc$mean[i]
    fc_lower_80[i] <- last_original[i] + fc$lower[i,1]
    fc_upper_80[i] <- last_original[i] + fc$upper[i,1]
  } else {
    fc_mean[i] <- fc_mean[i-24] + fc$mean[i]
    fc_lower_80[i] <- fc_lower_80[i-24] + fc$lower[i,1]
    fc_upper_80[i] <- fc_upper_80[i-24] + fc$upper[i,1]
  }
}

# Ensure non-negative
fc_mean <- pmax(fc_mean, 0)
fc_lower_80 <- pmax(fc_lower_80, 0)
fc_upper_80 <- pmax(fc_upper_80, 0)

# Create forecast timestamps
forecast_start <- max(combined_flow$Hour) + hours(1)
forecast_times <- seq(forecast_start, by = "hour", length.out = forecast_hours)

# Create results dataframe
forecast_results <- data.frame(
  Hour = forecast_times,
  Forecast_Flow = round(fc_mean, 2),
  Lower_80 = round(fc_lower_80, 2),
  Upper_80 = round(fc_upper_80, 2)
)

cat("\n========================================\n")
cat("FORECAST RESULTS (First 24 hours)\n")
cat("========================================\n")
print(head(forecast_results, 24))

```

-   **Forecast horizon:** 168 hours = 7 days × 24 hours
-   **Auto ARIMA:** Automatically selects best (p,d,q) parameters based
    on AIC
-   **ARIMA(3,0,3):** Chosen for this data - handles autocorrelation
    well
-   **AIC comparison:** Lower AIC indicates better model fit

## Visualize the Forecast

```{r}
# Plot forecast with historical data
ggplot() +
  # Historical data (last 7 days)
  geom_line(data = tail(combined_flow, 168), 
            aes(x = Hour, y = Total_Flow), 
            color = "black", size = 0.6) +
  # Forecast
  geom_line(data = forecast_results, 
            aes(x = Hour, y = Forecast_Flow), 
            color = "blue", size = 1) +
  # Confidence interval
  geom_ribbon(data = forecast_results, 
              aes(x = Hour, ymin = Lower_80, ymax = Upper_80), 
              fill = "blue", alpha = 0.2) +
  # Vertical line separating historical from forecast
  geom_vline(xintercept = max(combined_flow$Hour), 
             linetype = "dashed", color = "red") +
  labs(title = "Water Flow Forecast - Next 7 Days",
       subtitle = "ARIMA(3,0,3) Model | 80% Confidence Intervals",
       x = "Date/Time", 
       y = "Total Flow Rate") +
  scale_x_datetime(date_breaks = "1 day", date_labels = "%b %d") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

## Weekly Report

```{r}
# Summarize forecast by day
forecast_summary <- forecast_results %>%
  mutate(
    Date = as.Date(Hour),
    Hour_of_Day = hour(Hour)
  ) %>%
  group_by(Date) %>%
  summarise(
    Daily_Total = sum(Forecast_Flow),
    Daily_Avg = mean(Forecast_Flow),
    Peak_Hour = Hour_of_Day[which.max(Forecast_Flow)],
    Peak_Flow = max(Forecast_Flow),
    Min_Flow = min(Forecast_Flow),
    .groups = 'drop'
  )

cat("\n========================================\n")
cat("7-DAY FORECAST SUMMARY\n")
cat("========================================\n")
print(forecast_summary)

cat("\n========================================\n")
cat("TOTAL FORECAST\n")
cat("========================================\n")
cat("Total flow for next 7 days:", 
    format(round(sum(forecast_results$Forecast_Flow), 0), big.mark = ","), "\n")
cat("Average daily flow:", 
    format(round(mean(forecast_summary$Daily_Total), 0), big.mark = ","), "\n")

```

**The water flow data is stationary after differencing and successfully
forecast for one week forward.**

## Key Observations from forecast:

**Day 1 (Dec 3, 2015):**

-   **Lower total flow:** 257 (only partial day from 12:00 onward)

-   First forecast day starts at noon

**Days 2-7 (Dec 4-9, 2015):**

-   **Stable daily totals:** 1,060-1,065 per day
-   **Consistent average:** \~44 per hour
-   **Peak hour:** 2:00 AM consistently with 75-76 flow
-   **Min flow:** 5-7 during low periods

**Day 8 (Dec 10, 2015):**

-   **Partial day** (ends at 11:00 AM)

-   **Total:** 802 for first 11 hours

-   **Stable forecast:** Days 2-7 show consistent \~1,060 daily total

-   **Business insight:** Predictable patterns enable operational
    planning

**Real-World Business Value**

-   **Operational planning:** Schedule maintenance during low-flow
    periods
-   **Resource allocation:** Predict daily totals for capacity planning
-   **Risk management:** Use confidence intervals for buffer planning
-   **Pattern recognition:** Identifies peak usage times (2:00 AM)

This Project demonstrates the complete forecasting workflow from raw
data → cleaning → aggregation → stationarity testing → transformation →
model selection → forecasting → interpretation


------------------------------------------------------------------
